# CITE‑seq — Errors & Survival

## **Block 6 -  CITE-seq integration (ADT + RNA)** {#sec-block6}

::: content-box-gray
**Goal:** Bring the protein layer into the analysis. The RNA workflow up to **Block 5** ignored ADT entirely. The CITE-seq value-add is the ability to resolve dropout, validate annotation, and weight modalities per cell.
:::

```{r load_packages, echo=FALSE, message=FALSE, warning=FALSE}
pkgs <- c("Seurat", "SeuratObject", "ggplot2", "patchwork",
          "dplyr", "tidyr", "Matrix", "SingleR", 
          "harmony", "scales", "ggrepel", "BiocParallel",
          "scDblFinder", "SingleCellExperiment")

suppressPackageStartupMessages({
  library(Seurat);   library(SeuratObject); library(ggplot2)
  library(patchwork); library(dplyr);     library(tidyr);       library(Matrix)
  library(SingleR);   library(BiocParallel)
  library(harmony);  library(scales);       library(ggrepel)
})

#Instructor only
source("scripts/safe_accessor_function.R")
source("scripts/instructor_answers.R")
source("scripts/reveal_function.R")
```

Order of operations: inspect the ADT layer and contrast it with RNA, normalize, QC the panel, contrast RNA vs ADT for matched markers, gate, cross-check for swapped channels, build the ADT PCA, then integrate with WNN.

**Load file in R:**

```{r}
library(here) # install.packages("here")
sobj_filt <- readRDS(
  here("checkpoints", "outputs", "sobj_preprocessed.rds")
)
# top markers
top_markers <- readRDS(
  here("checkpoints", "outputs", "top_markers.rds")
)
```

```{r}
DefaultAssay(sobj_filt) <- "ADT"
adt_counts <- LayerData(sobj_filt, assay = "ADT", layer = "counts")

cat("ADT count matrix:\n")
cat("  Proteins (rows):", nrow(adt_counts), "\n")
cat("  Cells   (cols) :", ncol(adt_counts), "\n")
cat("  Class          :", class(adt_counts), "\n")
cat("  Layers in ADT  :", paste(SafeLayers(sobj_filt[["ADT"]]), collapse = ", "), "\n\n")
```

Per-protein summary: scale and distribution:

```{r}
adt_summary <- data.frame(
  protein  = rownames(adt_counts),
  min      = apply(adt_counts, 1, min),
  median   = apply(adt_counts, 1, median),
  mean     = round(Matrix::rowMeans(adt_counts), 2),
  max      = apply(adt_counts, 1, max),
  pct_zero = round(Matrix::rowMeans(adt_counts == 0) * 100, 1)
)
cat("Per-protein distribution (first 8 proteins):\n")
print(head(adt_summary, 8), row.names = FALSE)
```

Per-cell total ADT count

```{r}
total_adt_per_cell <- Matrix::colSums(adt_counts)
cat("\nPer-cell total ADT counts:\n")
print(round(quantile(total_adt_per_cell, c(0.05, 0.25, 0.5, 0.75, 0.95)), 0))

DefaultAssay(sobj_filt) <- "RNA"
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.0a 

An ADT protein with median ~0 but max in the high hundreds is typical bimodality (background cells vs stained cells). A protein with mean below 1 in every cell is something else. Which is which in the table above?

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Healthy bimodal: median near 0, mean a few units higher, max in the hundreds. The protein is unstained in most cells and strongly positive in its target subset. Dead antibody pattern: mean below 1, max also low (under 10), `pct_zero` very high. Nothing rises above background because the antibody is non-functional. In the injected dataset, CD56 shows the dead pattern; CD3, CD4, CD8a, CD14, CD19 show the bimodal pattern. Examine the table sorted by mean to spot the anomaly.

LIVE DEMO:

```{r}
DefaultAssay(sobj_filt) <- "ADT"
adt <- LayerData(sobj_filt, assay="ADT", layer="counts")
qc <- data.frame(
  protein  = rownames(adt),
  median   = apply(adt, 1, median),
  mean     = round(Matrix::rowMeans(adt), 2),
  max      = apply(adt, 1, max),
  pct_zero = round(Matrix::rowMeans(adt == 0)*100, 1)
)
qc[order(qc$mean), ]
DefaultAssay(sobj_filt) <- "RNA"
```
PLOT / OUTPUT: 

Could also use:

```{r}
RidgePlot(sobj_filt, features=rownames(sobj_filt[['ADT']])[1:6], assay='ADT')
```

:::

### Step 6.0b - RNA sparsity vs ADT sparsity

**Block 1** (@sec-block1) measured RNA sparsity on its own. Now that the ADT layer has been inspected too, the comparison is the point: the same cells, two assays, very different zero rates.

```{r}
rna_counts_b6 <- LayerData(sobj_filt, assay = "RNA", layer = "counts")
rna_sparsity  <- 1 - Matrix::nnzero(rna_counts_b6) / prod(dim(rna_counts_b6))
adt_sparsity  <- 1 - Matrix::nnzero(adt_counts)     / prod(dim(adt_counts))

cat("RNA sparsity:", round(rna_sparsity * 100, 1), "%\n")
cat("ADT sparsity:", round(adt_sparsity * 100, 1), "%\n")
```


:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.0b

Why is ADT sparsity so much lower than RNA sparsity? What does each zero mean biologically in the two modalities?
:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

RNA sparsity in this 500+9-gene subset is still substantial; ADT sparsity is typically below 5%. A single mRNA molecule must be captured by an oligo-dT primer, reverse-transcribed, amplified, and sequenced. The capture and RT steps fail at a substantial rate per molecule, producing a zero where the gene was expressed. This is dropout. An ADT zero comes from antibody staining: cells are incubated with hundreds-to-thousands of antibody molecules per protein, sequencing depth for protein tags is also higher per cell. The only way to read zero is for the protein to be absent or below detection background. RNA zeros mix true absence with capture failure; ADT zeros are mostly true absence.

LIVE DEMO:

```{r}
rna <- LayerData(sobj_filt, assay="RNA", layer="counts")
adt <- LayerData(sobj_filt, assay="ADT", layer="counts")
mean(rna == 0) * 100        # RNA sparsity %
mean(adt == 0) * 100        # ADT sparsity %
# CD4 specifically:
sum(rna["CD4", ] == 0) / ncol(rna) * 100
sum(adt["CD4", ] == 0) / ncol(adt) * 100
```
PLOT / OUTPUT: Console; histograms of *per-cell* zero rate also work for visual impact

:::

### Step 6.1 - Re-inspect the ADT layer

ADT has been sitting untouched since **Block 1** (@sec-block1). Confirm its state explicitly before normalizing.

```{r}
DefaultAssay(sobj_filt) <- "ADT"
cat("ADT assay current state:\n")
cat("  Active assay : ADT\n")
cat("  Proteins     :", nrow(sobj_filt[["ADT"]]), "\n")
cat("  Cells (post-QC):", ncol(sobj_filt), "\n")
cat("  Layers       :", paste(SafeLayers(sobj_filt[["ADT"]]), collapse = ", "), "\n")
```

At this point we expect: `counts only`. No `data`, no `scale.data`. Normalization is the next step.

:::: puzzle
::: {.callout-tip icon="false"}
## PUZZLE 6.1/A (ADT)

Without normalizing yet, which protein has the highest raw variance across cells? Is it a lineage marker (CD3, CD4, CD8a, CD14, CD19, CD56) or an activation marker (HLADR, CD69, CD25, PD1)?

```{r}
adt_counts <- LayerData(sobj_filt, assay = "ADT", layer = "counts")
adt_var <- apply(adt_counts, 1, var)
cat("\nTop 5 ADT proteins by raw-count variance:\n")
print(round(sort(adt_var, decreasing = TRUE)[1:5], 1))
```

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Lineage markers usually win: CD4, CD8a, CD14, CD19, CD3 have the highest raw variance because they go from background-low in non-target cells to very high in target cells. Activation markers like CD69, HLADR, CD25 have lower variance because they are expressed at moderate levels in many cells. If an activation marker tops the list, the dataset may be enriched for activated cells, or one antibody is behaving unusually.

LIVE DEMO:

```{r}
adt <- LayerData(sobj_filt, assay="ADT", layer="counts")
var_per_prot <- apply(adt, 1, var)
sort(var_per_prot, decreasing = TRUE)[1:5]
```
:::


:::: puzzle
::: {.callout-tip icon="false"}
## PUZZLE 6.1/B (ADT)

How many cells are positive for BOTH CD4 and CD8a in the `raw counts` (> 5 counts each)? In healthy PBMC this should be near zero; anything substantial is a doublet or a contamination signal.

```{r}
cd4_pos <- adt_counts["CD4", ] > 5
cd8_pos <- adt_counts["CD8a", ] > 5
cat("\nADT CD4+ cells              :", sum(cd4_pos), "\n")
cat("ADT CD8a+ cells             :", sum(cd8_pos), "\n")
cat("Double-positive (suspicious):", sum(cd4_pos & cd8_pos), "  (~0 expected)\n")
```


:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Double-positive CD4+/CD8a+ in raw ADT should be near zero in PBMC. More than 1-2% of cells: suspect doublets that scDblFinder missed, or antibody spillover. Investigate via nCount_RNA and total ADT for the suspect cells.

LIVE DEMO:

```{r}
cd4 <- adt["CD4", ] > 5
cd8 <- adt["CD8a", ] > 5
sum(cd4 & cd8)
table(cd4 & cd8, sobj_filt$is_dbl)
```
PLOT / OUTPUT: Console output

:::


:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.1

Does the double-positive count match what you expect biologically, or does it indicate doublets that survived scDblFinder? What would you do about it before annotating T cell subsets?

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Expected near zero (under 1% of cells). True double-positive CD4+/CD8+ cells are rare in PBMC (some MAIT and gamma-delta T cells). More than 1-2% and the cells are likely doublets that scDblFinder missed (homotypic T-cell doublets). Cross-check against scDblFinder doublet class; if many double-positives are not flagged as doublets, increase the scDblFinder threshold or remove the double-positives manually before annotating T-cell subsets.

LIVE DEMO:

```{r}
adt <- LayerData(sobj_filt, assay="ADT", layer="counts")
cd4 <- adt["CD4", ] > 5
cd8 <- adt["CD8a", ] > 5
table(double_pos = cd4 & cd8, scDbl = sobj_filt$is_dbl)
```

PLOT / OUTPUT: Console table; small counts in (TRUE, FALSE) cell are the suspicious cases

:::

### Step 6.2 - ADT normalization (CLR, margin = 2)

CLR (Centered Log-Ratio) normalization. The margin argument sets the
direction:
  + `margin = 1`: normalizes each protein across all cells (row means ~0)
  + `margin = 2`: normalizes each cell across all proteins (column means ~0)

`margin = 2` is correct for CITE-seq. It removes per-cell variation in total
antibody capture, analogous to library size normalization in RNA. `margin = 1`
runs without error but destroys biological signal. You will diagnose a real
object normalized the wrong way in **Block 8, Scenario 3**.

```{r}
DefaultAssay(sobj_filt) <- "ADT"

sobj_filt <- NormalizeData(
  sobj_filt,
  normalization.method = "CLR",
  margin               = 2  # CORRECT: normalizes each cell across proteins
)

adt_m2 <- LayerData(sobj_filt, layer = "data")
cat("Column means after margin=2 (per cell), expected ~0:\n")
print(round(colMeans(adt_m2)[1:8], 4))
cat("\nRow means (per protein), should NOT be ~0 (biological variation preserved):\n")
print(round(rowMeans(adt_m2), 4))

DefaultAssay(sobj_filt) <- "RNA"
cat("\nADT normalized correctly.\n")
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.2

Why is the row mean a useful diagnostic? What would the row means look like if margin had been set to 1?

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

After CLR with margin=2 (per cell across proteins), column means are ~0 by construction (each cell's protein values were centered). Row means (per protein across cells) carry the biological signal of which proteins are abundant and should NOT be near zero. They reflect dataset-wide protein abundance differences. If margin=1 was used instead (per protein across cells), row means would be near zero. That is the wrong-margin fingerprint. Cenario 3 of Block 7 contains exactly this failure.

LIVE DEMO:

```{r}
sobj_filt <- NormalizeData(sobj_filt, assay="ADT",
                           normalization.method="CLR", margin=2)
d <- LayerData(sobj_filt, assay="ADT", layer="data")
round(rowMeans(d), 3)
round(colMeans(d), 3)[1:5]
```

PLOT / OUTPUT: Console output; `row means != 0, col means ~ 0`.

:::

### Step 6.3 - ADT panel QC: detect dead / failed antibodies

A failed conjugation or degraded antibody produces a "dead channel": the protein reads near zero in every cell, with almost no variance. It does not error. If you gate or annotate on a dead channel, you silently lose a whole population.

```{r}
DefaultAssay(sobj_filt) <- "ADT"
adt_counts <- LayerData(sobj_filt, layer = "counts")

panel_qc <- data.frame(
  protein   = rownames(adt_counts),
  median    = round(apply(adt_counts, 1, median), 2),
  mean      = round(Matrix::rowMeans(adt_counts), 2),
  pct_zero  = round(Matrix::rowMeans(adt_counts == 0) * 100, 1),
  max       = apply(adt_counts, 1, max)
)
panel_qc <- panel_qc[order(panel_qc$mean), ]
cat("ADT panel QC (sorted by mean; suspicious = bottom rows):\n")
print(head(panel_qc, 6), row.names = FALSE)

DefaultAssay(sobj_filt) <- "RNA"
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.3

One protein has near-zero counts in essentially every cell while its RNA counterpart is clearly expressed in a defined cluster. Which protein, and which cell type does it mark?

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

In the injected dataset: CD56 (encoded by NCAM1). CD56 marks NK cells. With CD56 ADT dead, any NK-cell annotation based on ADT alone fails. The diagnostic chunk compares CD56 ADT (near-zero) with NCAM1 RNA (clear signal in the NK cluster) and surfaces the inconsistency. Habit: always cross-check the lowest-mean ADT proteins against their RNA counterparts.

LIVE DEMO:

```{r}
FeaturePlot(sobj_filt, c("CD56", "NCAM1"), reduction = "umap")
```

PLOT / OUTPUT: FeaturePlot panel: CD56 ADT (flat) next to NCAM1 RNA (positive in a cluster)

:::


Compare the suspect protein (ADT) against its RNA gene side by side.

Edit `suspect_adt` / `suspect_rna` to the protein flagged above.

```{r}
suspect_adt <- "CD56"     # protein flagged as near-zero in panel QC
suspect_rna <- "NCAM1"    # its RNA counterpart (NK marker)
```

Decide the title from the actual `panel_qc` numbers instead of asserting a fixed claim. On the clean object this protein is healthy; on the injected object it is dead. The title should say whichever is true for the object that is actually loaded.

```{r}
suspect_row <- panel_qc[panel_qc$protein == suspect_adt, ]
is_dead <- nrow(suspect_row) == 1 && suspect_row$pct_zero > 90 && suspect_row$mean < 1

adt_title <- if (is_dead) {
  paste0(suspect_adt, " ADT\n(flat: likely failed antibody)")
} else {
  paste0(suspect_adt, " ADT\n(signal present: not a dead channel here)")
}

DefaultAssay(sobj_filt) <- "ADT"
p_dead <- FeaturePlot(sobj_filt, suspect_adt, min.cutoff = "q05") +
  ggtitle(adt_title)

DefaultAssay(sobj_filt) <- "RNA"
p_alive <- FeaturePlot(sobj_filt, suspect_rna, min.cutoff = "q05") +
  ggtitle(paste0(suspect_rna, " RNA\n(clear NK signal: protein should exist)"))

p_dead | p_alive
DefaultAssay(sobj_filt) <- "RNA"
```

### Step 6.4 - RNA vs ADT for the same marker (dropout)

Direct comparison. CD4 mRNA reads as zero in many CD4+ T cells (dropout), while CD4 protein from ADT is bimodal across the same cells.

```{r}
DefaultAssay(sobj_filt) <- "RNA"
p_rna <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("CD4 via RNA\n(dropout: many CD4+ cells read as zero)") +
  scale_color_gradient(low = "lightgrey", high = "#991b1b")

DefaultAssay(sobj_filt) <- "ADT"
p_adt <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("CD4 via ADT protein\n(bimodal, reliable)") +
  scale_color_gradient(low = "lightgrey", high = "#0d7377")

p_rna | p_adt
DefaultAssay(sobj_filt) <- "RNA"
```

### Step 6.5 - RNA-protein correlation across all matched markers

For each ADT protein with an RNA counterpart, compute the **Spearman correlation** across all cells. Low correlation = high dropout in RNA.

```{r}
adt_rna_pairs <- list(
  CD3  = "CD3E",
  CD4  = "CD4",
  CD8a = "CD8A",
  CD14 = "CD14",
  CD19 = "CD19",
  CD56 = "NCAM1"
)

DefaultAssay(sobj_filt) <- "ADT"
adt_vals <- t(LayerData(sobj_filt, layer = "data"))
DefaultAssay(sobj_filt) <- "RNA"
rna_vals <- t(LayerData(sobj_filt, layer = "data"))

cat("Spearman correlation (RNA vs ADT) per marker:\n")
cat(sprintf("  %-8s %s\n", "Marker", "Spearman rho"))
cat(sprintf("  %-8s %s\n", "------", "------------"))

for (adt_name in names(adt_rna_pairs)) {
  rna_gene <- adt_rna_pairs[[adt_name]]
  if (rna_gene %in% colnames(rna_vals) &&
      adt_name %in% colnames(adt_vals)) {
    rho <- cor(rna_vals[, rna_gene],
               adt_vals[, adt_name],
               method = "spearman")
    cat(sprintf("  %-8s %.3f\n", adt_name, rho))
  }
}
DefaultAssay(sobj_filt) <- "RNA"
```

### Step 6.6 - Quantify dropout: ADT-positive but RNA-zero cells

For each marker, find cells that are clearly protein-positive (ADT > threshold) but have zero RNA counts. These are the cells that would be misannotated by RNA-only analysis.

```{r}
DefaultAssay(sobj_filt) <- "ADT"
adt_data <- LayerData(sobj_filt, layer = "data")
DefaultAssay(sobj_filt) <- "RNA"
rna_counts <- LayerData(sobj_filt, layer = "counts")

cat("Dropout analysis: cells positive by ADT but zero by RNA\n")
cat(sprintf("  %-8s %-12s %-12s %s\n",
            "Marker", "ADT+ cells", "RNA==0 among ADT+", "Dropout rate"))
cat(sprintf("  %-8s %-12s %-12s %s\n",
            "------", "----------", "-----------------", "------------"))

marker_pairs <- list(CD4 = "CD4", CD14 = "CD14", CD19 = "CD19")
for (adt_name in names(marker_pairs)) {
  rna_gene <- marker_pairs[[adt_name]]
  if (rna_gene %in% rownames(rna_counts) && adt_name %in% rownames(adt_data)) {
    adt_pos  <- which(adt_data[adt_name, ] > 1.0)
    rna_zero <- sum(rna_counts[rna_gene, adt_pos] == 0)
    pct      <- round(rna_zero / length(adt_pos) * 100, 1)
    cat(sprintf("  %-8s %-12d %-12d %s%%\n",
                adt_name, length(adt_pos), rna_zero, pct))
  }
}
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.6

For a marker with 80% dropout, what fraction of cells would be misannotated by an RNA-only workflow? Quick arithmetic:

```{r}
cat("\nWorked example: 80% dropout on CD4 RNA\n")
cat("  Single-marker decision (CD4 RNA > 0): misses 80% of true CD4+ cells.\n")
cat("  Marker panel of N independent markers (each with 80% dropout):\n")
cat("    P(all N drop)= 0.8^N\n")
for (n in c(1, 2, 3, 5)) {
  cat(sprintf("    N = %d markers -> %.1f%% of cells still misannotated\n",
              n, 0.8^n * 100))
}
cat("Rule of thumb: with 80% per-marker dropout, you need >= 3 independent\n")
cat("markers in the panel to reduce misannotation below 50%.\n")
```

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Single-marker decisions: up to 80% misannotated for cells whose RNA dropped out. The script prints the arithmetic for 1, 2, 3, 5 independent markers (each with 80% dropout): rule P(all drop) = 0.8^N. So 64%, 51%, 33% with 2, 3, 5 markers respectively. Annotation should never depend on a single marker for a high-dropout gene. ADT rescues because protein dropout is near zero.

LIVE DEMO:

```{r}
for (n in 1:5) cat(sprintf("N=%d -> %.1f%% still miss\n", n, 0.8^n*100))
```

PLOT / OUTPUT: Console output

:::

### Step 6.7 - Digital gating

Digital gating replicates the biaxial scatter plots from flow cytometry using ADT data. It allows computational cell classification with the same logic immunologists use at the bench.

```{r}
DefaultAssay(sobj_filt) <- "ADT"

color_by <- if ("singler_label_top" %in% colnames(sobj_filt@meta.data)) {
  "singler_label_top"
} else {
  "cell_type"
}

gate_data <- FetchData(sobj_filt, vars = c("CD4", "CD8a", color_by))
names(gate_data)[3] <- "label"

ggplot(gate_data, aes(CD4, CD8a, color = label)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_density_2d(color = "grey50", linewidth = 0.3) +
  geom_vline(xintercept = 1.5, linetype = "dashed", color = "#991b1b") +
  geom_hline(yintercept = 1.5, linetype = "dashed", color = "#991b1b") +
  annotate("text", x = 3.5, y = 0.4, label = "CD4+ T cells", size = 3.5) +
  annotate("text", x = 0.4, y = 3.5, label = "CD8+ T cells", size = 3.5) +
  labs(title    = "Digital gating: CD4 vs CD8a (ADT protein)",
       subtitle = "Only possible with ADT; RNA dropout makes this scatter uninformative",
       x = "CD4 (CLR normalized)", y = "CD8a (CLR normalized)",
       color = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.text = element_text(size = 7))

DefaultAssay(sobj_filt) <- "RNA"
```

### Step 6.7b - Silent failure: DefaultAssay forgotten between calls

A function call that "works" on the wrong assay produces a silent error. 

- `FeaturePlot("CD4")` with `DefaultAssay = "RNA"` plots the CD4 RNA gene.
- `FeaturePlot("CD4")` with `DefaultAssay = "ADT"` plots the CD4 protein.

The plot RENDERS in both cases, and the two plots can look deceptively similar: `min.cutoff = "q05"` rescales each panel's color gradient to its own value range, so a sparse RNA signal and a dense ADT signal both end up stretched across a similar-looking color scale. The failure is silent precisely because nothing in the rendered plot warns you which assay produced it.

Quantify the difference that the plot alone hides: what fraction of cells read exactly zero in each modality for this marker.

```{r}
DefaultAssay(sobj_filt) <- "RNA"
cd4_rna_zero_pct <- mean(LayerData(sobj_filt, layer = "counts")["CD4", ] == 0) * 100
DefaultAssay(sobj_filt) <- "ADT"
cd4_adt_zero_pct <- mean(LayerData(sobj_filt, layer = "counts")["CD4", ] == 0) * 100
DefaultAssay(sobj_filt) <- "RNA"

cat(sprintf("CD4 zero rate: %.1f%% of cells in RNA, %.1f%% of cells in ADT\n",
            cd4_rna_zero_pct, cd4_adt_zero_pct))
cat("The plots below can look similar at a glance; the zero rates above are\n")
cat("the actual difference the two assays are reporting for the same gene.\n")
```

Demonstrate by drawing the same call under both assays. Titles are kept short and font size is constrained, since the two plots render side by side at half width each.

```{r}
DefaultAssay(sobj_filt) <- "RNA"
p_as_rna <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("'CD4' | DefaultAssay = RNA") +
  theme(plot.title = element_text(size = 11))

DefaultAssay(sobj_filt) <- "ADT"
p_as_adt <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("'CD4' | DefaultAssay = ADT") +
  theme(plot.title = element_text(size = 11))

p_as_rna | p_as_adt
DefaultAssay(sobj_filt) <- "RNA"
```
:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.7b 

Both plots rendered, and at a glance they can look almost the same. The zero-rate numbers printed above tell a different story. If you saved one of these plots as a figure for a paper without checking `DefaultAssay` first, which one would you have, and would a reader be able to tell from the figure alone that 'CD4' meant something different in each panel?

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

You would have whichever assay was active at the moment of the plot call, and a reader could not tell which one from the figure alone: the plot title says only 'CD4', the axes are identical UMAP coordinates, and `min.cutoff='q05'` rescales each panel's own color gradient independently, so a sparse RNA signal and a denser ADT signal end up stretched across visually similar-looking scales. This is exactly why the zero-rate numbers matter: they show the real difference in what fraction of cells read as positive in each modality, a difference the rescaled color gradient hides. The published figure could silently show CD4 RNA when CD4 protein was intended, or vice versa, and nothing in the image itself would flag the mismatch. Habit: set `DefaultAssay` explicitly at the top of any plotting block, AND/OR pass the assay inside the call; better still, edit the plot title to include the assay name and consider reporting the zero rate alongside the figure so the modality difference is documented, not just visible to whoever already knows to look for it.

LIVE DEMO:

```{r}
DefaultAssay(sobj_filt) <- "RNA"
mean(LayerData(sobj_filt, layer="counts")["CD4", ] == 0) * 100   # RNA zero rate
DefaultAssay(sobj_filt) <- "ADT"
mean(LayerData(sobj_filt, layer="counts")["CD4", ] == 0) * 100   # ADT zero rate
DefaultAssay(sobj_filt) <- "RNA"
FeaturePlot(sobj_filt, "CD4") + ggtitle("CD4 protein (ADT)")
FeaturePlot(sobj_filt, "CD4") + ggtitle("CD4 mRNA (RNA)")
```

PLOT / OUTPUT: Two `FeaturePlots` side by side, one per assay, distinct titles; the zero-rate numbers explain what the plots alone do not show.

:::

### Step 6.7c - Spot the bug

Read the code below. It looks correct. What is wrong?

```{r}
DefaultAssay(sobj_filt) <- "RNA"
adt_markers <- FindMarkers(sobj_filt, ident.1 = "CD4Tcell",
                             group.by = "cell_type",
                             min.pct  = 0.25)
```

Your answer:

**REVEAL:** the call asks for markers of CD4 T cells but DefaultAssay is RNA.
It returns RNA markers, not ADT markers. The output is correct for RNA but
does not answer the question "which proteins distinguish CD4 T cells".
Either set `DefaultAssay = "ADT"` first, or pass `assay = "ADT"` inside the
`FindMarkers` call. The `min.pct = 0.25` is also wrong for ADT (designed for
sparse RNA); for ADT use a value like 0.5 or 0.0 with `logfc.threshold`.

### Step 6.8 - When gating makes no biological sense: a swapped channel

A swapped or mislabeled channel runs perfectly. CLR normalizes it, `FeaturePlot` draws it, gating gates it. Nothing errors. The only signal is that the biology is impossible. The defense is to cross-check each protein against an independent label (its RNA marker).

Each protein should correlate with its own RNA gene across cells. 

A canonical marker with rho near 0 is a mislabeled channel.

```{r}
DefaultAssay(sobj_filt) <- "ADT"
adt_d <- t(LayerData(sobj_filt, layer = "data"))
DefaultAssay(sobj_filt) <- "RNA"
rna_d <- t(LayerData(sobj_filt, layer = "data"))

check_pairs <- list(CD4 = "CD4", CD8a = "CD8A", CD19 = "CD19",
                    CD14 = "CD14", CD3 = "CD3E")

cat(sprintf("  %-6s %-12s %-10s\n", "ADT", "rho(self)", "verdict"))
cat(sprintf("  %-6s %-12s %-10s\n", "---", "---------", "-------"))
for (a in names(check_pairs)) {
  g <- check_pairs[[a]]
  if (a %in% colnames(adt_d) && g %in% colnames(rna_d)) {
    rho <- round(cor(adt_d[, a], rna_d[, g], method = "spearman"), 3)
    verdict <- if (rho < 0.05) "SUSPECT" else "ok"
    cat(sprintf("  %-6s %-12s %-10s\n", a, rho, verdict))
  }
}
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.8: 

Two ADT channels show rho near 0 against their own RNA gene, while the cross pair is high. Confirm the swap with the visual below.

```{r}
DefaultAssay(sobj_filt) <- "ADT"
g <- FetchData(sobj_filt, vars = c("CD14", "CD19"))
DefaultAssay(sobj_filt) <- "RNA"
g$CD19_rna <- FetchData(sobj_filt, vars = "CD19")[, 1]  # B cell RNA marker

ggplot(g, aes(CD14, CD19, color = CD19_rna)) +
  geom_point(alpha = 0.5, size = 0.8) +
  scale_color_gradient(low = "lightgrey", high = "#991b1b") +
  labs(title    = "CD14 vs CD19 (ADT), colored by CD19 RNA",
       subtitle = "If CD19 RNA piles up on the CD14 axis, the labels are swapped",
       x = "CD14 (ADT)", y = "CD19 (ADT)", color = "CD19 RNA") +
  theme_classic(base_size = 12)
DefaultAssay(sobj_filt) <- "RNA"
```
:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Spearman rho between each ADT protein and its RNA counterpart across cells. Healthy markers show rho 0.3-0.7. Swap fingerprint: ADT_X vs RNA_X near zero AND ADT_X vs RNA_Y high; same with X,Y reversed. In the injected dataset CD14 and CD19 ADT rownames are swapped. The fix block rebuilds the ADT assay with rownames re-swapped and re-runs CLR. Conditional, so on clean data it does nothing.

LIVE DEMO:

```{r}
DefaultAssay(sobj_filt) <- "ADT"
FeatureScatter(sobj_filt, "CD14", "CD19") + ggtitle("ADT CD14 vs ADT CD19")
DefaultAssay(sobj_filt) <- "RNA"
FeatureScatter(sobj_filt, "CD14", "CD19") + ggtitle("RNA CD14 vs RNA CD19")
```

PLOT / OUTPUT: Two scatter plots side by side. Healthy: ADT and RNA scatters match. With a swap: they diverge

:::

- `v5-safe`: rebuild the ADT assay from its count matrix with corrected names.
- `Conditional`: only acts if the swap is actually detected (safe on clean data).

```{r}
DefaultAssay(sobj_filt) <- "ADT"
adt_d <- t(LayerData(sobj_filt, layer = "data"))
rna_d <- t(LayerData(sobj_filt, assay = "RNA", layer = "data"))

swapped <- FALSE
if (all(c("CD14", "CD19") %in% colnames(adt_d))) {
  rho14 <- cor(adt_d[, "CD14"], rna_d[, "CD14"], method = "spearman")
  swapped <- is.na(rho14) || rho14 < 0.05
}

if (swapped) {
  adt_counts <- LayerData(sobj_filt, assay = "ADT", layer = "counts")
  rn  <- rownames(adt_counts)
  i14 <- which(rn == "CD14"); i19 <- which(rn == "CD19")
  rn[c(i14, i19)] <- rn[c(i19, i14)]
  rownames(adt_counts) <- rn
  # CreateAssay5Object() builds a fresh assay with only the counts layer, so
  # the [[<- replacement method warns that it differs in structure (no
  # data/scale.data layers yet) from the ADT assay it replaces. Expected and
  # harmless here; suppressWarnings() wraps the assignment itself, since the
  # notice comes from the replacement method, not from CreateAssay5Object().
  new_adt_assay <- CreateAssay5Object(counts = adt_counts)
  suppressWarnings(sobj_filt[["ADT"]] <- new_adt_assay)
  sobj_filt <- NormalizeData(sobj_filt, normalization.method = "CLR",
                             margin = 2, verbose = FALSE)
  cat("ADT labels corrected (CD14 <-> CD19) and re-normalized.\n")
} else {
  cat("No swap detected; ADT labels left unchanged.\n")
}
```

```{r}
# Re-check
adt_d <- t(LayerData(sobj_filt, layer = "data"))
rna_d <- t(LayerData(sobj_filt, assay = "RNA", layer = "data"))
for (a in c("CD14", "CD19")) {
  rho <- round(cor(adt_d[, a], rna_d[, a], method = "spearman"), 3)
  cat(sprintf("  %-6s rho(self) = %s\n", a, rho))
}
DefaultAssay(sobj_filt) <- "RNA"
```

### Step 6.9 - WNN: compute ADT PCA, then integrate

Weighted Nearest Neighbor integration learns the relative contribution of each modality per cell. Cells where RNA is informative receive more RNA weight; cells where protein is more discriminating receive more ADT weight. WNN requires a PCA per modality. RNA PCA exists from Block 4; we now build the ADT PCA.

```{r}
DefaultAssay(sobj_filt) <- "ADT"

adt_features <- rownames(sobj_filt[["ADT"]])  # all 24 proteins

sobj_filt <- ScaleData(
  sobj_filt,
  features = adt_features,
  verbose  = FALSE
)
```

With 24 proteins, requesting 20 PCs triggers a Seurat warning about computing too many singular values via truncated SVD. Two fixes both work:

- (a) reduce npcs to 15 (still captures the structure), or
- (b) pass approx = FALSE to use exact SVD.

We use (a) because 15 PCs is enough signal for 24 proteins.

```{r}
sobj_filt <- RunPCA(
  sobj_filt,
  features       = adt_features,
  npcs           = 15,
  reduction.name = "pca.adt",
  reduction.key  = "pcaADT_",
  verbose        = FALSE
)

cat("Reductions now available:", paste(SafeReductions(sobj_filt), collapse = ", "), "\n")
```

`Seurat` expects one weight name per modality. Passing only "RNA.weight" triggers a warning that ADT.weight has been auto-assigned. Pass both explicitly to silence the notice and document the intended column names.

```{r}
sobj_filt <- FindMultiModalNeighbors(
  sobj_filt,
  reduction.list       = list("pca", "pca.adt"),
  dims.list            = list(1:20, 1:15),
  modality.weight.name = c("RNA.weight", "ADT.weight"),
  verbose              = FALSE
)

sobj_filt <- RunUMAP(
  sobj_filt,
  nn.name        = "weighted.nn",
  reduction.name = "wnn.umap",
  reduction.key  = "wnnUMAP_",
  verbose        = FALSE
)

DefaultAssay(sobj_filt) <- "RNA"
cat("WNN complete. New reduction: wnn.umap\n")
```

### Step 6.10 - RNA-only UMAP vs WNN UMAP

WNN does not necessarily move every cell type to a visually different region; UMAP layouts can rotate or mirror between runs even when the underlying neighborhoods barely change. The two panels below can look similar at first glance. What actually matters is whether the set of cells grouped together changed, not whether the picture looks different. 

**Step 6.11c** (later) checks this directly per cell-type via a cross-table; here, a quick clustering comparison gives a first read before that.

```{r}
sobj_filt <- FindClusters(sobj_filt, resolution = 0.5,
                          cluster.name = "rna_only_clusters_preview",
                          verbose = FALSE)
rna_only_clusters <- sobj_filt$rna_only_clusters_preview

sobj_filt <- FindClusters(sobj_filt, graph.name = "wsnn", resolution = 0.5,
                          cluster.name = "wnn_clusters_preview",
                          verbose = FALSE)
wnn_clusters_preview <- sobj_filt$wnn_clusters_preview

agreement_tab <- table(RNA_only = rna_only_clusters, WNN = wnn_clusters_preview)
cat("Cluster membership: RNA-only vs WNN (rows: RNA-only clusters, cols: WNN clusters)\n")
print(agreement_tab)
```

A simple agreement score: for each RNA-only cluster, what fraction of its cells land in the single WNN cluster that captures the most of them.

```{r}
best_match_frac <- apply(agreement_tab, 1, function(row) max(row) / sum(row))
cat(sprintf("\nMedian per-cluster agreement: %.1f%% of cells stay grouped together\n",
            100 * median(best_match_frac)))

p_rna <- DimPlot(sobj_filt, reduction = "umap",
                 group.by = "singler_label_top",
                 label = TRUE, label.size = 2.5, repel = TRUE) +
  ggtitle("RNA-only UMAP") + NoLegend()

p_wnn <- DimPlot(sobj_filt, reduction = "wnn.umap",
                 group.by = "singler_label_top",
                 label = TRUE, label.size = 2.5, repel = TRUE) +
  ggtitle("WNN UMAP (RNA + ADT)") + NoLegend()

p_rna | p_wnn
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.10

If the median agreement above is high (cells mostly stay grouped the same way) but the two UMAP layouts still look visually different (different rotation, different relative positions), what does that tell you about reading UMAP plots side by side versus comparing cluster membership directly?

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

It means the visual comparison and the cluster-membership comparison are answering different questions, and only one of them is actually about whether WNN changed the result. UMAP is a 2D projection chosen independently each time it is run; rotation, mirroring, and relative spacing between blobs can differ between two UMAP runs even on the exact same underlying neighborhood graph, because UMAP's layout optimization has no fixed orientation to anchor to. High agreement in the cluster cross-table means the cells that were grouped together under RNA-only are still grouped together under WNN, which is the substantive claim. A side-by-side UMAP comparison is useful for a first visual impression and for spotting gross differences (a population that splits or merges), but it is not a reliable way to judge subtle changes, and two UMAPs that look different are not, by themselves, evidence that WNN changed anything about which cells belong together.

LIVE DEMO:

```{r}
print(agreement_tab)
median(best_match_frac)   # fraction of cells staying grouped together
```

PLOT / OUTPUT: The agreement_tab cross-table and the median agreement number, alongside the two UMAP panels

:::

### Step 6.11 - Per-cell modality weights

- `RNA.weight` close to 1 = cell is better characterized by RNA
- `RNA.weight` close to 0 = cell is better characterized by ADT
- `RNA.weight` and ADT.weight sum to 1 for each cell (`FindMultiModalNeighbors` normalizes them that way), so 0.5 is the natural reference point: below it, ADT is contributing more than RNA to that cell's placement in the WNN graph, not just "some amount" of protein information.

```{r}
pct_below_half <- round(mean(sobj_filt$RNA.weight < 0.5) * 100, 1)
cat(sprintf("Cells where ADT contributes more than RNA (RNA.weight < 0.5): %.1f%%\n",
            pct_below_half))

ggplot(sobj_filt@meta.data, aes(x = RNA.weight, fill = singler_label_top)) +
  geom_histogram(bins = 40, alpha = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "#2c3142") +
  facet_wrap(~singler_label_top, scales = "free_y") +
  labs(title    = "Per-cell RNA modality weight from WNN",
       subtitle = "Dashed line at 0.5: left of it, ADT outweighs RNA for that cell",
       x = "RNA weight (1 = fully RNA, 0 = fully ADT)") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none",
        strip.text      = element_text(size = 7))
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.11 

Which cell types in this dataset rely most on ADT (low RNA weight)? Why does that match what you know about RNA dropout for their defining markers?

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

CD4 T cells and CD8 T cells, by a wide margin. Their defining markers (CD4, CD8A) have the worst RNA dropout. B cells (MS4A1, CD79A) and monocytes (CD14, LYZ) have stronger RNA signal and rely less on ADT. NK cells (NKG7, GNLY, NCAM1) are intermediate. The histogram facet of RNA.weight by cell type makes this concrete, and the dashed line at RNA.weight = 0.5 marks the point where ADT outweighs RNA for that cell, not just contributes some amount of information.

LIVE DEMO:

```{r}
mean(sobj_filt$RNA.weight < 0.5) * 100   # cells where ADT outweighs RNA
ggplot(sobj_filt@meta.data, aes(x = RNA.weight, fill = singler_label_top)) +
  geom_histogram(bins = 40) + geom_vline(xintercept = 0.5, linetype = "dashed") +
  facet_wrap(~singler_label_top)
```

PLOT / OUTPUT: Histogram of RNA.weight faceted by cell type, with a dashed line at 0.5. T cells peak left of the line, monocytes peak right of it. 

:::

### Step 6.11b - RNA.weight extremes per cluster: are any clusters protein-driven?

A cluster where most cells have RNA.weight near 0 is identified almost entirely by protein. That can be biology (T cell subsets, where RNA dropout is severe) or an artifact (a protein-specific batch effect concentrated in one cluster).

```{r}
rna_w_by_cluster <- sobj_filt@meta.data %>%
  group_by(seurat_clusters) %>%
  summarise(median_rna_w = median(RNA.weight),
            mean_rna_w   = mean(RNA.weight),
            n_cells      = n(),
            .groups = "drop") %>%
  arrange(median_rna_w)

cat("RNA.weight per cluster (sorted; low = protein-driven):\n")
print(rna_w_by_cluster)
```

Flag clusters where median RNA.weight < 0.3 (protein dominates)

```{r}
protein_driven <- rna_w_by_cluster$seurat_clusters[rna_w_by_cluster$median_rna_w < 0.3]
if (length(protein_driven) > 0) {
  cat("\nClusters where protein dominates (median RNA.weight < 0.3):\n")
  cat("  ", paste(protein_driven, collapse = ", "), "\n")
  cat("Inspect these clusters: are they T-cell subsets (expected) or",
      "something else?\n")
}
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.11b 

For the most protein-driven cluster, what fraction of its cells were assigned to a T-cell label by SingleR? The block below computes the answer and applies the decision rule.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Computed in the script. Decision thresholds: >=70% T-cell -> expected RNA-dropout behavior, no action; 30-70% -> mixed, inspect ADT proteins driving the dominance; <30% -> technical artifact, check ADT batch, CLR margin, isotype background. The script applies the rule automatically and prints the verdict.

LIVE DEMO:

```{r}
md <- sobj_filt@meta.data
cs <- aggregate(RNA.weight ~ seurat_clusters, data = md, FUN = median)
cs[order(cs$RNA.weight), ]
```

PLOT / OUTPUT: Console table.
:::

T-cell-like SingleR labels (any case)

```{r}
t_cell_pattern <- "T[._ -]?cell|CD4|CD8|Tcell|T cell"
```

Most protein-driven `cluster = smallest median RNA.weight`

```{r}
md <- sobj_filt@meta.data
cluster_summary <- md %>%
  group_by(seurat_clusters) %>%
  summarise(median_rna_w = median(RNA.weight, na.rm = TRUE),
            n_cells      = n(),
            t_cell_frac  = mean(grepl(t_cell_pattern, singler_label,
                                      ignore.case = TRUE), na.rm = TRUE),
            .groups = "drop") %>%
  arrange(median_rna_w)

cat("\nClusters sorted by median RNA.weight (lowest = most protein-driven):\n")
print(cluster_summary)

target_cluster <- cluster_summary$seurat_clusters[1]
tfrac          <- cluster_summary$t_cell_frac[1]
mrw            <- cluster_summary$median_rna_w[1]

cat(sprintf("\nMost protein-driven cluster: %s (median RNA.weight = %.2f)\n",
            target_cluster, mrw))
cat(sprintf("Fraction of its cells with T-cell label: %.1f%%\n", tfrac * 100))

cat("\nDecision rule (concrete thresholds):\n")
cat("  t_cell_frac >= 70%  -> protein dominance reflects RNA dropout (expected,\n")
cat("                         CD4/CD8 RNA drops out heavily). No action.\n")

cat("  t_cell_frac 30-70%  -> mixed. Inspect ADT panel for this cluster: is one\n")
cat("                         specific protein driving the dominance? Run:\n")
cat("                         FeaturePlot(sobj_filt, '<protein>',\n")
cat("                                     cells = WhichCells(sobj_filt, idents = '<cluster>'))\n")
cat("  t_cell_frac < 30%   -> technical explanation. Check:\n")
cat("                         1. ADT batch (Block 6.3, panel QC)\n")
cat("                         2. CLR margin (Block 6.2; verify row means != 0)\n")
cat("                         3. Isotype background (Block 8 Scenario 4)\n")

verdict <- if (tfrac >= 0.70) {
  "RNA dropout (expected)"
} else if (tfrac >= 0.30) {
  "mixed; inspect ADT panel"
} else {
  "technical artifact; check batch/CLR/isotype"
}
cat(sprintf("\nVerdict for cluster %s: %s\n", target_cluster, verdict))
```

### Step 6.11c - Cell-level reannotation: WNN clustering vs RNA-only label

Run a graph-based clustering on the WNN graph and compare it to the RNA-only SingleR annotation, restricted to cells where ADT dominated.

```{r}
sobj_filt <- FindClusters(sobj_filt,
                          graph.name = "wsnn",
                          resolution = 0.5,
                          verbose    = FALSE)
sobj_filt$wnn_clusters <- Idents(sobj_filt)
```

Cells where ADT dominated WNN (RNA.weight < 0.3)

```{r}
low_rna_mask <- sobj_filt$RNA.weight < 0.3
n_low <- sum(low_rna_mask)
cat("Cells where ADT dominated WNN (RNA.weight < 0.3):", n_low,
    sprintf("(%.1f%% of total)\n", 100 * n_low / ncol(sobj_filt)))

if (n_low >= 20) {
  cat("\nCross-tab: RNA-based SingleR label vs WNN cluster (ADT-dominated cells)\n")
  print(table(
    SingleR_label = sobj_filt$singler_label[low_rna_mask],
    WNN_cluster   = sobj_filt$wnn_clusters[low_rna_mask]
  ))
}
```


:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.11c

For the row of the cross-table with the largest disagreement, the SingleR-RNA label and the WNN cluster diverge. Which one do you trust? The code below identifies that row automatically and runs the deciding check.

```{r}
if (n_low >= 20) {
  ctab <- table(
    SingleR_label = sobj_filt$singler_label[low_rna_mask],
    WNN_cluster   = sobj_filt$wnn_clusters[low_rna_mask]
  )
  # Largest off-diagonal cell = biggest disagreement
  if (length(ctab) > 1) {
    max_cell <- arrayInd(which.max(ctab), dim(ctab))
    src_lbl <- rownames(ctab)[max_cell[1]]
    src_clu <- colnames(ctab)[max_cell[2]]
    n_conf  <- ctab[max_cell[1], max_cell[2]]
    cat(sprintf("\nLargest disagreement: %d cells labeled '%s' by SingleR-RNA\n",
                n_conf, src_lbl))
    cat(sprintf("but assigned to WNN cluster %s.\n", src_clu))

    # Deciding check: marker expression in those cells.
    # T-cell markers (RNA + ADT), B-cell markers, Mono markers
    decider_markers <- list(
      T_cell    = list(rna = c("CD3E", "CD3D"), adt = c("CD3")),
      B_cell    = list(rna = c("MS4A1", "CD79A"), adt = c("CD19", "CD20")),
      Monocyte  = list(rna = c("CD14", "LYZ"),   adt = c("CD14")),
      NK_cell   = list(rna = c("NKG7", "GNLY"),  adt = c("CD56"))
    )

    conf_cells <- which(low_rna_mask &
                        sobj_filt$singler_label == src_lbl &
                        sobj_filt$wnn_clusters  == src_clu)
    cat(sprintf("Marker detection in %d disputed cells:\n", length(conf_cells)))
    for (ct in names(decider_markers)) {
      for (mk in decider_markers[[ct]]$rna) {
        if (mk %in% rownames(sobj_filt[["RNA"]])) {
          v <- FetchData(sobj_filt, vars = mk)[conf_cells, 1]
          cat(sprintf("  %s (RNA %s): %.1f%% > 0\n", ct, mk, mean(v > 0) * 100))
        }
      }
      for (mk in decider_markers[[ct]]$adt) {
        if (mk %in% rownames(sobj_filt[["ADT"]])) {
          DefaultAssay(sobj_filt) <- "ADT"
          v <- FetchData(sobj_filt, vars = mk)[conf_cells, 1]
          DefaultAssay(sobj_filt) <- "RNA"
          cat(sprintf("  %s (ADT %s): %.1f%% > 1.0 (CLR units)\n",
                      ct, mk, mean(v > 1.0) * 100))
        }
      }
    }

    cat("\nDecision rule:\n")
    cat("  The cell type whose markers fire in >= 50% of cells (RNA + ADT\n")
    cat("  combined) wins. If two cell types tie, mark these cells as\n")
    cat("  ambiguous (likely doublets that scDblFinder missed) and remove\n")
    cat("  before downstream group comparison.\n")
  }
}
```


:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Decision rule from the script: cross-check marker expression in the disputed cells. The cell type whose markers (RNA + ADT) fire in >=50% of cells wins. If two cell types tie, mark the cells as ambiguous (likely undetected doublets) and exclude them from downstream group comparisons. Do NOT pick the WNN label by default; do NOT pick the SingleR label by default. Let the markers adjudicate.

LIVE DEMO:

```{r}
low_rna_mask <- sobj_filt$RNA.weight < 0.3
tab <- table(sobj_filt$singler_label[low_rna_mask],
             sobj_filt$wnn_clusters[low_rna_mask])
tab
```

PLOT / OUTPUT: Console cross-table.

:::

### Step 6.12 - Condition comparison: Healthy vs COVID-19

This plot looks informative. But with 3 Healthy and 4 COVID-19 donors, it can be completely driven by one outlier donor.

```{r}
group_props <- sobj_filt@meta.data %>%
  filter(!is.na(singler_label_top), !is.na(condition)) %>%
  group_by(condition, singler_label_top) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(condition) %>%
  mutate(prop = n / sum(n))

ggplot(group_props, aes(x = condition, y = prop, fill = singler_label_top)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_y_continuous(labels = percent_format()) +
  labs(title    = "Cell type proportions by condition",
       subtitle = "Looks clear at the group level; check whether one donor drives it",
       x = "Condition", y = "Proportion", fill = "Cell type") +
  theme_classic(base_size = 12) +
  theme(legend.text = element_text(size = 8))

donor_props <- sobj_filt@meta.data %>%
  filter(!is.na(singler_label_top)) %>%
  group_by(donor_id, condition, singler_label_top) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(donor_id) %>%
  mutate(prop = n / sum(n))

ggplot(donor_props, aes(x = donor_id, y = prop, fill = singler_label_top)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~condition, scales = "free_x") +
  labs(title    = "Cell type proportions per donor",
       subtitle = "Consistent pattern across donors within each group; not driven by one outlier",
       x = NULL, y = "Proportion (%)", fill = "Cell type") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.text = element_text(size = 7),
        strip.text  = element_text(face = "bold"))
```

### Step 6.12b - Statistical reality check: donor-level test with power note {#sec-step6.12b}

Per-donor proportions for ONE example cell type, tested between conditions. Run this for whichever label your annotation produced as the largest in COVID-19; the script tries to pick automatically.

Per-donor proportions for ONE example cell type, tested between conditions. Run this for whichever label your annotation produced as the largest in COVID-19; the script tries to pick automatically. Restricted to top-N labels: a label held by 1-2 cells could show a spurious 100% delta between conditions by chance alone, which would make a meaningless "finding" look like the strongest signal in the dataset.

```{r}
per_donor <- sobj_filt@meta.data %>%
  filter(!is.na(condition), !is.na(singler_label_top), condition %in% c("Healthy", "COVID19")) %>%
  group_by(donor_id, condition, singler_label_top) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(donor_id, condition) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()
```

Pick the cell type with the largest delta between groups for demonstration

```{r}
delta_by_lbl <- per_donor %>%
  group_by(singler_label_top, condition) %>%
  summarise(mean_prop = mean(prop), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = condition, values_from = mean_prop) %>%
  mutate(delta = abs(COVID19 - Healthy)) %>%
  arrange(desc(delta))

target_lbl <- delta_by_lbl$singler_label_top[1]
cat("Largest mean-proportion delta is for label:", as.character(target_lbl), "\n")
cat(sprintf("  Healthy mean: %.1f%%   COVID-19 mean: %.1f%%   delta: %.1f%%\n",
            100 * delta_by_lbl$Healthy[1],
            100 * delta_by_lbl$COVID19[1],
            100 * delta_by_lbl$delta[1]))

subset_df <- per_donor %>% filter(singler_label_top == target_lbl)
n_h <- sum(subset_df$condition == "Healthy")
n_c <- sum(subset_df$condition == "COVID19")

if (n_h >= 2 && n_c >= 2) {
  wt <- wilcox.test(prop ~ condition, data = subset_df, exact = FALSE)
  cat(sprintf("\nWilcoxon rank-sum (n=%d Healthy vs n=%d COVID-19):\n", n_h, n_c))
  cat(sprintf("  W = %g, p = %.3f\n", wt$statistic, wt$p.value))
}
```

Power floor: with `n=3 vs n=4` and `Wilcoxon` at `alpha=0.05`, the smallest detectable standardized effect size (Cohen's d) is roughly 1.5 for 80% power.

**In plain terms:** only proportion differences larger than ~1.5 standard deviations of the donor-level distribution are detectable. Any `p-value` here is exploratory, not confirmatory.

```{r}
cat("\n--- Power note ---\n")
cat("n=3 vs n=4 donors. Minimum detectable effect size (Cohen's d) ~1.5 for\n")
cat("80% power at alpha=0.05. Treat any p-value here as exploratory only.\n")
```

:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.12b

Assuming the printed p-value is in the 0.05 - 0.10 range, what conclusion can you DRAW from this dataset? The block below applies the explicit decision rule for small-sample group comparison.

```{r}
cat("\nDecision matrix for the observed p-value:\n")
if (exists("wt")) {
  pv <- wt$p.value
  cat(sprintf("  Observed p-value: %.3f\n", pv))
  cat("\n  p < 0.01 with n=7 donors total: STRONG signal, replicate in a\n")
  cat("                                     larger cohort before claiming.\n")
  cat("  0.01 <= p < 0.05               : Suggestive. Report effect size +\n")
  cat("                                     per-donor plot. Not confirmatory.\n")
  cat("  0.05 <= p < 0.10               : Exploratory. Cannot claim difference.\n")
  cat("                                     State the n, the direction, and the\n")
  cat("                                     effect size; flag as hypothesis-\n")
  cat("                                     generating only.\n")
  cat("  p >= 0.10                      : No evidence of difference at this n.\n")
  cat("                                     Power analysis says n>=10 per group\n")
  cat("                                     needed for the observed effect size.\n")

  # Auto-apply
  verdict <- if (pv < 0.01) {
    "STRONG (replicate in larger cohort)"
  } else if (pv < 0.05) {
    "suggestive (report with caveats)"
  } else if (pv < 0.10) {
    "exploratory (hypothesis-generating only)"
  } else {
    "no evidence of difference at this n"
  }
  cat(sprintf("\nAutomatic verdict at p = %.3f: %s\n", pv, verdict))

  # Minimum n needed for the observed effect (Cohen's d -> n via power calc)
  pooled_sd <- sd(subset_df$prop)
  obs_d     <- delta_by_lbl$delta[1] / pooled_sd
  cat(sprintf("\nObserved Cohen's d (donor-level): %.2f\n", obs_d))
  cat("Approximate n per group needed for 80%% power at alpha=0.05:\n")
  # Rough rule: n ~= 16 / d^2 for two-sample t-test, similar for Wilcoxon
  if (!is.na(obs_d) && obs_d > 0) {
    n_needed <- ceiling(16 / obs_d^2)
    cat(sprintf("  ~%d donors per group (using rule n ~ 16/d^2)\n", n_needed))
  }
}

saveRDS(sobj_filt, "outputs/sobj_annotated.rds")
cat("Annotated object saved: outputs/sobj_annotated.rds\n")
```


:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Exploratory only. With n=7 donors total, a p-value in this range is hypothesis-generating and cannot support a published claim of difference. Report: n per group, direction of the difference, effect size, statement that confirmatory analysis requires a larger cohort. The script estimates the n needed for 80% power at the observed effect size (rule of thumb n ~ 16 / d^2).

LIVE DEMO: See **Step 6.12b** (@sec-step6.12b) in the script: it auto-applies the decision matrix.

PLOT / OUTPUT: Console output with automatic verdict.

:::


:::: question
::: {.callout-tip icon="false"}
## QUESTION 6.12

How would you formally test the group-level difference given only 3 healthy vs 4 COVID-19 donors? What is the minimum reporting standard you would accept in a paper?

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

Donor as the unit of analysis. Compute per-donor cell-type proportions (one number per donor per cell type), Wilcoxon rank-sum across donors. With n=3 vs n=4 power is minimal: only effect sizes near Cohen's d=1.5 are detectable at alpha=0.05. Minimum reporting standard: plot per-donor proportions (Step 6.12 does this), state the n, and explicitly mark the analysis as exploratory unless effect sizes are very large. Group-level barplots without per-donor backing should not appear in a paper.

LIVE DEMO:

```{r}
subset_df <- per_donor %>% filter(singler_label == target_lbl)
wilcox.test(prop ~ condition, data = subset_df, exact = FALSE)
print(subset_df[, c("donor_id", "condition", "prop")])
```
PLOT / OUTPUT: Per-donor proportion boxplot with individual donors as points, colored by condition.

:::


## **Block 7 - Closing** {#sec-block7}

::: content-box-gray
Final view of the annotated proportions and the R session record. The pitfalls reference table lives in the instructor's guide PDF.
::: 

### Step 7.1 - Four-panel summary

Four results that together summarize the course, reusing objects already built in Blocks 5 and 6 rather than recomputing anything:
  
1. Final WNN UMAP with confident annotation: the end state of the RNA + ADT integration.
2. Top canonical markers per cell type: the evidence behind the labels in panel 1.
3. Cell type proportions by condition: the biological comparison the whole pipeline was built to support.
4. CD4 in RNA vs ADT: the single clearest illustration from the course of why the protein layer earns its place in the analysis.
     
```{r}
p_final_umap <- DimPlot(sobj_filt, reduction = "wnn.umap",
                        group.by = "singler_label_clean",
                        label = TRUE, label.size = 2.5, repel = TRUE) +
  ggtitle("1. Final WNN UMAP (confident annotation)") + NoLegend()

p_final_markers <- DotPlot(sobj_filt,
                           features = unique(top_markers$gene),
                           idents   = setdiff(levels(Idents(sobj_filt)), "Ambiguous"),
                           group.by = "singler_label_clean") +
  RotatedAxis() +
  theme(axis.text.x = element_text(size = 7), legend.position = "none") +
  ggtitle("2. Canonical markers behind the labels")

if ("singler_label_top" %in% colnames(sobj_filt@meta.data)) {
  final_props <- sobj_filt@meta.data %>%
    filter(!is.na(condition), !is.na(singler_label_top)) %>%
    group_by(condition, singler_label_top) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(condition) %>%
    mutate(prop = n / sum(n))

  p_final_props <- ggplot(final_props, aes(condition, prop, fill = singler_label_top)) +
    geom_bar(stat = "identity", width = 0.65) +
    scale_y_continuous(labels = percent_format()) +
    labs(x = "Condition", y = "Proportion", fill = NULL) +
    theme_classic(base_size = 10) +
    theme(legend.text = element_text(size = 6),
          legend.position = "right",
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle("3. Cell type proportions by condition")
}

DefaultAssay(sobj_filt) <- "RNA"
p_cd4_rna <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("4a. CD4 (RNA)") +
  theme(plot.title = element_text(size = 10), legend.position = "none")
DefaultAssay(sobj_filt) <- "ADT"
p_cd4_adt <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("4b. CD4 (ADT)") +
  theme(plot.title = element_text(size = 10), legend.position = "none")
DefaultAssay(sobj_filt) <- "RNA"
p_final_cd4 <- p_cd4_rna | p_cd4_adt

(p_final_umap | p_final_markers) / (p_final_props | p_final_cd4)
```
     
### Step 7.2 - Session info

```{r}
sessionInfo()
```

