# CITE‑seq — Errors & Survival

## **Block 6** {#sec-block6}

::: content-box-gray

::: 


:::: question
::: {.callout-tip icon="false"}
## QUESTION 2.8

Is your dataset's percent.mt distribution closer to a fraction or a percentage? Run `summary(sobj$percent.mt)` to confirm and remember this trap when reading code from other groups.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER



LIVE DEMO:

```{r}

```
:::


:::: question
::: {.callout-tip icon="false"}
## QUESTION 2.8

Is your dataset's percent.mt distribution closer to a fraction or a percentage? Run `summary(sobj$percent.mt)` to confirm and remember this trap when reading code from other groups.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

LIVE DEMO:

```{r}

```
:::

## **Block 7** {#sec-block7}

::: content-box-gray

::: 


:::: question
::: {.callout-tip icon="false"}
## QUESTION 2.8

Is your dataset's percent.mt distribution closer to a fraction or a percentage? Run `summary(sobj$percent.mt)` to confirm and remember this trap when reading code from other groups.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER



LIVE DEMO:

```{r}

```
:::


:::: question
::: {.callout-tip icon="false"}
## QUESTION 2.8

Is your dataset's percent.mt distribution closer to a fraction or a percentage? Run `summary(sobj$percent.mt)` to confirm and remember this trap when reading code from other groups.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

LIVE DEMO:

```{r}

```
:::

