# Extra

## **Block 8**

::: content-box-gray
**Goal:** compute QC metrics, look at their distributions across donors, and filter cells using thresholds chosen from the data, not from a tutorial.
:::


:::: question
::: {.callout-tip icon="false"}
## QUESTION 2.8

Is your dataset's percent.mt distribution closer to a fraction or a percentage? Run `summary(sobj$percent.mt)` to confirm and remember this trap when reading code from other groups.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER



LIVE DEMO:

```{r}

```
:::


:::: question
::: {.callout-tip icon="false"}
## QUESTION 2.8

Is your dataset's percent.mt distribution closer to a fraction or a percentage? Run `summary(sobj$percent.mt)` to confirm and remember this trap when reading code from other groups.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

LIVE DEMO:

```{r}

```
:::

## **Block 9**

::: content-box-gray
**Goal:** compute QC metrics, look at their distributions across donors, and filter cells using thresholds chosen from the data, not from a tutorial.
:::



:::: question
::: {.callout-tip icon="false"}
## QUESTION 2.8

Is your dataset's percent.mt distribution closer to a fraction or a percentage? Run `summary(sobj$percent.mt)` to confirm and remember this trap when reading code from other groups.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER



LIVE DEMO:

```{r}

```
:::


:::: question
::: {.callout-tip icon="false"}
## QUESTION 2.8

Is your dataset's percent.mt distribution closer to a fraction or a percentage? Run `summary(sobj$percent.mt)` to confirm and remember this trap when reading code from other groups.

:::
::::

::: {.callout-tip icon="false" collapse="true"}
### ANSWER

LIVE DEMO:

```{r}

```
:::