---
title: "Single-Cell Survival Course"
subtitle: "CITE-seq Troubleshooting with Seurat v5"
author: "Workshop | GSE149689 | RNA + ADT | 7 donors | 3,000 cells"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: false
    self_contained: true
    code_folding: show
    df_print: paged
    css: course_style.css
---

```{r global_setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 11,
  fig.height = 5,
  fig.align  = "center",
  comment    = "#>"
)
```

------------------------------------------------------------------------

# Single-Cell Survival Course

**CITE-seq Troubleshooting with Seurat v5**

`GSE149689` \| `RNA + ADT` \| `7 donors` \| `3,000 cells` \| `24 proteins` \| `Wilk et al., Nature Medicine 2021`

------------------------------------------------------------------------

> **How to use this notebook** Run every chunk in order from top to bottom. Some chunks will produce errors. That is intentional. When an error or suspicious result appears, read the paragraph below the chunk before running the next one. Sections marked **Challenge** ask you to answer a question before continuing. Sections marked **Survival moment** flag decisions that ruin analyses silently.

**Pre-course checklist**

- [ ] R 4.3 or higher installed
- [ ] RStudio ready and a new session started (Session \> Restart R)
- [ ] Packages installed (Block 0, Step 0.1)
- [ ] Dataset downloaded to `data/gse149689_subset_3k.rds`
- [ ] `sobj_broken.rds` kept private until Block 5

------------------------------------------------------------------------

------------------------------------------------------------------------

# ══════════════════════════════════════════

# BLOCK 0 \| Setup

# ══════════════════════════════════════════

*Run at home before the course*

------------------------------------------------------------------------

## Step 0.1 \| Install packages

This chunk checks every package before installing. If `install.packages()` fails, it retries automatically via BiocManager. Safe to run multiple times.

```{r install_auto}
# [BLOCK 0 | Step 0.1 | Package installation with CRAN + BiocManager fallback]

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

all_pkgs <- c(
  "Seurat", "SeuratObject", "ggplot2", "patchwork",
  "dplyr", "Matrix", "harmony", "scales", "ggrepel",
  "SingleR", "celldex", "BiocParallel", "scDblFinder"
)

for (pkg in all_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message("Installing: ", pkg)
    tryCatch(install.packages(pkg, quiet = TRUE),
             error   = function(e) NULL,
             warning = function(w) NULL)
    if (!requireNamespace(pkg, quietly = TRUE)) {
      message("  CRAN failed. Retrying via BiocManager...")
      tryCatch(
        BiocManager::install(pkg, ask = FALSE, update = FALSE),
        error   = function(e) message("  FAILED: ", e$message),
        warning = function(w) NULL
      )
    }
    if (requireNamespace(pkg, quietly = TRUE))
      message("  OK: ", pkg, " (", as.character(packageVersion(pkg)), ")")
    else
      message("  FAILED: ", pkg, " -- run the fallback chunk below")
  } else {
    message("Already installed: ", pkg,
            " (", as.character(packageVersion(pkg)), ")")
  }
}
```

## Step 0.1b \| Fallback: force reinstall individual packages

Only run lines for packages that appeared as FAILED above. After running any line here, do **Session \> Restart R** before continuing.

```{r install_fallback, eval=FALSE}
# [BLOCK 0 | Step 0.1b | Force reinstall via BiocManager]
BiocManager::install("Seurat",       force = TRUE)
BiocManager::install("SeuratObject", force = TRUE)
BiocManager::install("ggplot2",      force = TRUE)
BiocManager::install("patchwork",    force = TRUE)
BiocManager::install("dplyr",        force = TRUE)
BiocManager::install("Matrix",       force = TRUE)
BiocManager::install("harmony",      force = TRUE)
BiocManager::install("scales",       force = TRUE)
BiocManager::install("ggrepel",      force = TRUE)
BiocManager::install("SingleR",      force = TRUE)
BiocManager::install("celldex",      force = TRUE)
BiocManager::install("BiocParallel", force = TRUE)
BiocManager::install("scDblFinder",  force = TRUE)
```

## Step 0.2 \| Verify and load all libraries

All packages are loaded here. No `library()` calls appear later in the notebook.

> If you see `namespace 'ggplot2' is imported by 'Seurat'...`, that is not an error. The package is already active. Always do **Session \> Restart R** before opening the notebook to avoid namespace conflicts.

```{r load_libs}
# [BLOCK 0 | Step 0.2 | Verify and load all libraries]

pkgs <- c("Seurat", "SeuratObject", "ggplot2", "patchwork",
          "dplyr", "Matrix", "SingleR", "celldex",
          "harmony", "scales", "ggrepel", "BiocParallel")

for (pkg in pkgs)
  cat(sprintf("  %-18s %s\n", pkg,
              ifelse(requireNamespace(pkg, quietly = TRUE), "OK", "MISSING")))

cat("\nSeurat version:", as.character(packageVersion("Seurat")), "\n")

suppressPackageStartupMessages({
  library(Seurat);   library(SeuratObject); library(ggplot2)
  library(patchwork); library(dplyr);       library(Matrix)
  library(SingleR);  library(celldex);      library(BiocParallel)
  library(harmony);  library(scales);       library(ggrepel)
})
cat("\nAll libraries loaded.\n")
```

## Step 0.3 \| Create folders and download dataset

```{r folders}
# [BLOCK 0 | Step 0.3 | Folder structure]
dir.create("data",    showWarnings = FALSE)
dir.create("outputs", showWarnings = FALSE)
cat("Folders ready.\n")
```

```{r download, eval=FALSE}
# [BLOCK 0 | Step 0.3 | Download dataset from Google Drive]
# FILE_ID is set to the course dataset. Change only if you have a different file.
FILE_ID <- "1mCg4VD91qK7pg-zL1iYUUBFWBr7fmmCM"
options(timeout = 600)
download.file(
  url      = paste0("https://drive.google.com/uc?export=download&confirm=t&id=",
                    FILE_ID),
  destfile = "data/gse149689_subset_3k.rds",
  mode     = "wb"
)
cat("File exists:", file.exists("data/gse149689_subset_3k.rds"), "\n")
cat("Size (MB)  :", round(file.size("data/gse149689_subset_3k.rds") / 1e6, 1), "\n")
```

------------------------------------------------------------------------

------------------------------------------------------------------------

# ══════════════════════════════════════════

# BLOCK 1 \| Foundations and Dataset

# ══════════════════════════════════════════

*0:00 to 0:30 \| 30 min*

------------------------------------------------------------------------

## Step 1.1 \| Why single-cell RNA-seq?

Bulk RNA-seq collapses an entire sample into one number per gene: the population average. A signal that appears in 10% of cells at high intensity and in 90% at zero will look like a dim signal in every cell. You cannot tell which cells express it, or whether the expressing cells are a specific subpopulation.

scRNA-seq gives each cell its own profile. The key trade-off:

| Property                            | Bulk RNA-seq       | scRNA-seq      |
|-------------------------------------|--------------------|----------------|
| Resolution                          | Population average | Per-cell       |
| Sensitivity for low-expressed genes | High               | Low (dropout)  |
| Sparsity                            | Dense matrix       | \~90% zeros    |
| Throughput (cells)                  | All                | Sampled subset |
| Cost per sample                     | Low                | High           |

The \~90% sparsity is not instrument noise. A transcript detected at low copy number will simply not be captured in many cells even if truly expressed. This is called **dropout**, and it drives most of the analytical decisions in scRNA-seq: normalization, dimensionality reduction, and annotation.

```{r bulk_vs_sc, fig.height=5}
# [BLOCK 1 | Step 1.1 | Simulated bulk vs single-cell comparison]
set.seed(42)
n  <- 200
sc <- rbind(
  data.frame(CD3E = rnorm(n/2, 8, 1), CD14 = rnorm(n/2, 1, 0.5),
             Type = "T cell"),
  data.frame(CD3E = rnorm(n/2, 1, 0.5), CD14 = rnorm(n/2, 8, 1),
             Type = "Monocyte")
)
bulk <- data.frame(CD3E = mean(sc$CD3E), CD14 = mean(sc$CD14))

ggplot(sc, aes(CD3E, CD14, color = Type)) +
  geom_point(alpha = 0.55, size = 1.8) +
  geom_point(data = bulk, color = "#111827", size = 9, shape = 18) +
  annotate("text", x = bulk$CD3E + 0.4, y = bulk$CD14,
           label = "Bulk average\n(no real cell)", hjust = 0, size = 3.2) +
  scale_color_manual(values = c("T cell" = "#0d7377", "Monocyte" = "#92400e")) +
  labs(title    = "Bulk RNA-seq loses the structure",
       subtitle = "The average is a ghost: it corresponds to no real biological state",
       x = "CD3E expression (T cell marker)",
       y = "CD14 expression (Monocyte marker)") +
  theme_classic(base_size = 13)
```

## Step 1.2 \| CITE-seq: adding the protein layer

CITE-seq (Cellular Indexing of Transcriptomes and Epitopes by Sequencing) extends scRNA-seq by simultaneously measuring surface protein abundance. Antibodies conjugated to short DNA barcodes (Antibody-Derived Tags, ADT) are stained on cells, then sequenced alongside the RNA in the same library.

Each cell yields two measurement vectors:

| Modality | Features | Typical values | Main strength |
|----|----|----|----|
| RNA | \~500 genes (full: \~33,000) | 0-50 UMIs per gene | Transcriptional state, pathways, trajectories |
| ADT | 24 surface proteins | 1-1,000 counts per protein | Clean bimodal signal, gating, protein abundance |

The protein layer is critical for three reasons:

1.  **Dropout**: canonical surface markers like CD4, CD8, CD14 are frequently undetected at the RNA level even in cells that clearly express the protein.
2.  **Gating**: protein data replicates the information from flow cytometry directly, connecting computational results to established biology.
3.  **Integration**: the WNN (Weighted Nearest Neighbor) algorithm can combine both modalities, giving each cell a per-modality weight based on information content.

## Step 1.3 \| Dataset: GSE149689

**Wilk et al., Nature Medicine 2021** : CITE-seq of PBMCs from:

- 3 healthy donors (Donor01, Donor02, Donor03)
- 4 COVID-19 donors: Donor04 (Mild), Donor05 (Moderate), Donor06 (Severe), Donor07 (Severe)

Cell types represented: CD4 T, CD8 T, NK, classical Monocyte, non-classical Monocyte, B cell, Plasmablast, DC, Platelet.

We use a synthetic 3,000-cell subset that mirrors the biological structure of the full dataset: same donor proportions, same cell type composition, same ADT panel, and an interferon response signal in COVID-19 donors.

------------------------------------------------------------------------

> **Discussion pause**
>
> - CD4 protein cleanly separates CD4 T cells by flow cytometry. Why does CD4 RNA often fail to do the same in scRNA-seq?
> - Plasmablasts are expanded in severe COVID-19. What RNA markers would you expect to be elevated? What ADT markers?
> - What is the minimum number of donors per condition to make a group-level claim statistically defensible?

------------------------------------------------------------------------

------------------------------------------------------------------------

# ══════════════════════════════════════════

# BLOCK 2 \| Object Inspection and QC

# ══════════════════════════════════════════

*0:30 to 1:00 \| 30 min*

------------------------------------------------------------------------

## Step 2.1 \| Load the object

```{r load_object}
# [BLOCK 2 | Step 2.1 | Load Seurat object]
sobj <- readRDS("data/gse149689_subset_3k.rds")
print(sobj)
```

## Step 2.2 \| Object anatomy: Seurat v4 vs v5

Seurat v5 changed how count matrices are stored internally. Scripts written for v4 break in v5 in ways that are not always obvious. Run the next two chunks exactly as they are.

```{r v4_slot_direct, error=TRUE}
# [BLOCK 2 | Step 2.2 | v4 direct slot access -- will error in v5]
# In Seurat v4, this was the standard way to get the count matrix.
counts_old <- sobj@assays$RNA@counts
```

The `@counts` slot no longer exists in Seurat v5 `Assay5` objects. The count matrix lives inside a layer system.

```{r v4_getassaydata, error=TRUE}
# [BLOCK 2 | Step 2.2 | GetAssayData with slot= argument -- deprecated in v5]
counts_old2 <- GetAssayData(sobj, slot = "counts")
```

`slot=` is deprecated. The v5 argument is `layer=`. Both of these errors are extremely common when receiving objects from collaborators or running old pipeline scripts.

```{r v5_correct_access}
# [BLOCK 2 | Step 2.2 | Correct v5 access patterns]

# Pattern 1: LayerData() -- recommended for v5
counts_rna <- LayerData(sobj, assay = "RNA", layer = "counts")
cat("RNA counts matrix dimensions (genes x cells):\n")
cat(" ", dim(counts_rna), "\n\n")

# Pattern 2: assay class
cat("RNA assay class :", class(sobj[["RNA"]]), "\n")
cat("ADT assay class :", class(sobj[["ADT"]]), "\n\n")

# Pattern 3: available layers
cat("RNA layers      :", paste(Layers(sobj[["RNA"]]), collapse = ", "), "\n")
cat("ADT layers      :", paste(Layers(sobj[["ADT"]]), collapse = ", "), "\n\n")

# Pattern 4: Seurat object version
cat("Object version  :", as.character(sobj@version), "\n")
```

## Step 2.3 \| Deep object inspection (intermediate challenges)

These are not tutorial exercises. Each question requires understanding what is inside the Seurat object and how to get it out.

```{r challenge_sparsity}
# [BLOCK 2 | Step 2.3 | Challenge 1: sparsity of the count matrix]
# Sparsity = fraction of zero entries. Drives normalization decisions.

counts_mat <- LayerData(sobj, assay = "RNA", layer = "counts")
total_entries  <- prod(dim(counts_mat))
nonzero        <- Matrix::nnzero(counts_mat)
sparsity       <- 1 - nonzero / total_entries

cat("RNA count matrix:\n")
cat("  Genes   :", nrow(counts_mat), "\n")
cat("  Cells   :", ncol(counts_mat), "\n")
cat("  Total entries  :", format(total_entries, big.mark = ","), "\n")
cat("  Non-zero entries:", format(nonzero, big.mark = ","), "\n")
cat("  Sparsity        :", round(sparsity * 100, 1), "%\n\n")

# ADT sparsity for comparison
adt_counts  <- LayerData(sobj, assay = "ADT", layer = "counts")
adt_nonzero <- Matrix::nnzero(adt_counts)
adt_total   <- prod(dim(adt_counts))
cat("ADT count matrix:\n")
cat("  Proteins:", nrow(adt_counts), "\n")
cat("  Cells   :", ncol(adt_counts), "\n")
cat("  Sparsity:", round((1 - adt_nonzero / adt_total) * 100, 1), "%\n")
```

> **Challenge 1:** Why is ADT sparsity so much lower than RNA sparsity? What does that imply about the biological interpretation of zero values in each modality?

```{r challenge_memory}
# [BLOCK 2 | Step 2.3 | Challenge 2: memory footprint of layers]
# Dense vs sparse storage matters at scale. A full 10x experiment with
# 50,000 cells and 33,000 genes as a dense matrix exceeds 50 GB.

# Compare sparse (counts) vs dense (after scaling) memory
counts_rna_sz <- object.size(LayerData(sobj, assay = "RNA", layer = "counts"))
cat("RNA counts layer (sparse):", format(counts_rna_sz, units = "Mb"), "\n")

# The full object
obj_size <- object.size(sobj)
cat("Full Seurat object       :", format(obj_size, units = "Mb"), "\n\n")

cat("Key insight: scale.data stores a DENSE matrix (genes x cells).\n")
cat("After ScaleData(), memory footprint increases substantially.\n")
cat("In full datasets: only scale the genes you will use in PCA.\n")
```

```{r challenge_pca_embeddings}
# [BLOCK 2 | Step 2.3 | Challenge 3: navigating reductions and embeddings]
# After RunPCA(), where do the PC coordinates live?
# How do you get them for a specific cell?

# First, check what reductions exist right now (before PCA has been run)
cat("Reductions in the object:\n")
print(Reductions(sobj))
cat("(empty -- no PCA yet, that comes in Block 3)\n\n")

# After PCA runs, the embeddings will be at:
# sobj@reductions$pca@cell.embeddings  -- cells x dims matrix
# Embeddings(sobj, reduction = "pca")  -- same thing, recommended
# sobj@reductions$pca@feature.loadings -- gene loadings

# For now, check the metadata in detail
cat("Metadata columns and types:\n")
str(sobj@meta.data)
```

```{r challenge_metadata_query}
# [BLOCK 2 | Step 2.3 | Challenge 4: querying cells by metadata]
# Real analysis constantly requires subsetting or querying cells by
# combinations of metadata. Practice the patterns here.

# How many cells per donor?
cat("Cells per donor:\n")
print(table(sobj$donor_id))

# How many cells per condition?
cat("\nCells per condition:\n")
print(table(sobj$condition))

# Cells from COVID-19 donors with more than 100 genes detected
covid_highgene <- sobj@meta.data %>%
  filter(condition == "COVID19", nFeature_RNA > 100)
cat("\nCOVID-19 cells with nFeature_RNA > 100:", nrow(covid_highgene), "\n")

# Cross-tabulation: condition x severity
cat("\nCondition x Severity:\n")
print(table(sobj$condition, sobj$severity))
```

```{r challenge_adt_names}
# [BLOCK 2 | Step 2.3 | Challenge 5: ADT feature naming and access]
# ADT feature names often differ from RNA gene names.
# CD3 protein != CD3E gene. CD8a protein != CD8A gene.
# This mismatch is a frequent source of confusion.

cat("ADT protein names:\n")
print(rownames(sobj[["ADT"]]))

cat("\nRNA marker names for comparison:\n")
rna_markers <- c("CD3E", "CD4", "CD8A", "CD14", "CD19", "NCAM1")
cat(rna_markers, "\n")

cat("\nImportant: 'CD8a' in ADT vs 'CD8A' in RNA.\n")
cat("FetchData() handles this transparently, but FeaturePlot()\n")
cat("requires you to be on the correct DefaultAssay first.\n")

# Demonstrate: what happens when you try to access an ADT feature
# using the RNA gene name
cat("\nDoes 'CD8A' exist in ADT?\n")
cat("'CD8A' in ADT rownames:", "CD8A" %in% rownames(sobj[["ADT"]]), "\n")
cat("'CD8a' in ADT rownames:", "CD8a" %in% rownames(sobj[["ADT"]]), "\n")
```

------------------------------------------------------------------------

> **Discussion pause**
>
> - The counts layer is sparse; the scale.data layer is dense. What does this mean for large datasets (50,000 cells, 33,000 genes)?
> - You receive an object where `Reductions(sobj)` returns `"pca"` and `"umap"` but `Layers(sobj[["RNA"]])` returns only `"counts"`. The data layer is missing. Can you trust the UMAP? What would you do?
> - What is the difference between `sobj[["RNA"]]` and `sobj@assays[["RNA"]]` in Seurat v5?

------------------------------------------------------------------------

## Step 2.4 \| Quality control metrics

### 2.4.1 \| Calculate QC metrics

Mitochondrial percentage (`percent.mt`) measures the fraction of UMIs from mitochondrial genes. High values indicate either stressed/dying cells (cytoplasmic RNA degrades faster than mitochondrial) or low-quality droplets. Ribosomal percentage (`percent.ribo`) measures ribosomal protein gene expression, which can be used to flag highly proliferating or metabolically active cells.

```{r qc_metrics}
# [BLOCK 2 | Step 2.4.1 | QC metrics]
sobj[["percent.mt"]]   <- PercentageFeatureSet(sobj, pattern = "^MT-")
sobj[["percent.ribo"]] <- PercentageFeatureSet(sobj, pattern = "^RP[SL]")

summary(sobj@meta.data[, c("nFeature_RNA", "nCount_RNA", "percent.mt")])
```

### 2.4.2 \| Visualize distributions

```{r qc_violin, fig.height=5}
# [BLOCK 2 | Step 2.4.2 | QC violin plots]
VlnPlot(sobj,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3, pt.size = 0.05, alpha = 0.3)
```

```{r qc_scatter, fig.width=12, fig.height=4}
# [BLOCK 2 | Step 2.4.2 | QC scatter plots]
p1 <- FeatureScatter(sobj, "nCount_RNA", "nFeature_RNA") +
  ggtitle("Counts vs genes detected") +
  theme(plot.title = element_text(size = 10))
p2 <- FeatureScatter(sobj, "nCount_RNA", "percent.mt") +
  ggtitle("Counts vs mitochondrial %") +
  theme(plot.title = element_text(size = 10))
p1 | p2
```

> **Challenge 6:** In the scatter plot of nCount_RNA vs nFeature_RNA, cells that fall far above the main diagonal are likely doublets (two cells captured together). Cells that cluster at the lower left are likely empty droplets. Do you see either population in this dataset? Write code to count cells where nFeature_RNA \> 200 AND nCount_RNA \< 500.

```{r challenge_outlier_detection}
# [BLOCK 2 | Step 2.4.2 | Challenge 6: outlier detection]
# Cells that are high in nFeature relative to nCount might be doublets.
# Low nCount with low nFeature = empty droplet.
# High percent.mt alone = stressed/dying cell.

# Cells that are potential outliers: nFeature > 2 SD above mean
mean_feat <- mean(sobj$nFeature_RNA)
sd_feat   <- sd(sobj$nFeature_RNA)
high_feat <- sum(sobj$nFeature_RNA > mean_feat + 2 * sd_feat)
cat("Cells with nFeature > mean + 2 SD (potential doublets):",
    high_feat, "\n")

# Cells with high mt: > 95th percentile
mt_95 <- quantile(sobj$percent.mt, 0.95)
high_mt <- sum(sobj$percent.mt > mt_95)
cat("Cells with percent.mt > 95th percentile:", high_mt, "\n")

# Distribution per condition: is mt% elevated in COVID-19?
cat("\nMedian percent.mt by condition:\n")
sobj@meta.data %>%
  group_by(condition) %>%
  summarise(median_mt = round(median(percent.mt), 2),
            mean_mt   = round(mean(percent.mt),   2),
            .groups   = "drop") %>%
  print()
```

### 2.4.3 \| Applying published thresholds without looking at the data

The original GSE149689 paper used these thresholds on the full 10x dataset with \~33,000 genes. We apply them to our 500-gene dataset without checking.

```{r filter_wrong, error=TRUE}
# [BLOCK 2 | Step 2.4.3 | Applying tutorial thresholds blindly -- ERROR]
sobj_filt <- subset(
  sobj,
  subset = nFeature_RNA > 200 &
           nFeature_RNA < 5000 &
           percent.mt   < 20
)
```

**"No cells found"** is the canonical result of copying thresholds from a tutorial without checking your own data distribution. The threshold `nFeature_RNA > 200` was calibrated for a transcriptome of \~33,000 genes where cells typically detect 2,000-3,000 genes. In a 500-gene panel, most cells detect 50-150 genes. The filter removes every single cell.

```{r thresholds_comparison}
# [BLOCK 2 | Step 2.4.3 | Compare published vs data-driven thresholds]
cat("--- Published thresholds (full 10x, 33k genes) ---\n")
cat("  nFeature_RNA > 200   removes empty droplets\n")
cat("  nFeature_RNA < 5000  removes likely doublets\n")
cat("  percent.mt   < 20%   removes dead/stressed cells\n\n")

cat("--- This dataset (500 genes) ---\n")
print(round(quantile(sobj$nFeature_RNA,
      probs = c(0.01, 0.05, 0.25, 0.5, 0.75, 0.95, 0.99))))

cat("\n--- Data-driven thresholds ---\n")
min_features <- max(10, round(quantile(sobj$nFeature_RNA, 0.02)))
max_features <-          round(quantile(sobj$nFeature_RNA, 0.98))
max_mt       <-          round(quantile(sobj$percent.mt,   0.95), 1)

cat("  nFeature_RNA > ", min_features, "  (same intent: empty droplets)\n")
cat("  nFeature_RNA < ", max_features, " (same intent: doublets)\n")
cat("  percent.mt   < ", max_mt, "%  (same intent: dead cells)\n")
```

```{r filter_correct}
# [BLOCK 2 | Step 2.4.3 | Apply data-driven thresholds]
cat("Cells before filtering:", ncol(sobj), "\n")

sobj_filt <- subset(
  sobj,
  subset = nFeature_RNA > min_features &
           nFeature_RNA < max_features &
           percent.mt   < max_mt
)

cat("Cells after filtering :", ncol(sobj_filt), "\n")
cat("Cells removed         :", ncol(sobj) - ncol(sobj_filt), "\n")
cat("Retention rate        :",
    round(ncol(sobj_filt) / ncol(sobj) * 100, 1), "%\n")

if (ncol(sobj_filt) < 500) {
  cat("\nWARNING: fewer than 500 cells remaining.\n")
  cat("Options: relax max_mt or lower min_features,\n")
  cat("or request pre-filtered matrices from GEO.\n")
} else {
  cat("\nCell count adequate for downstream analysis.\n")
}
```

------------------------------------------------------------------------

------------------------------------------------------------------------

# ══════════════════════════════════════════

# BLOCK 3 \| Preprocessing

# ══════════════════════════════════════════

*1:00 to 1:55 \| 55 min*

------------------------------------------------------------------------

## Step 3.1 \| Normalization

### 3.1.1 \| RNA normalization

LogNormalize divides each cell's counts by its total counts, multiplies by a scale factor (10,000 by default), then applies log1p. This removes the effect of variable sequencing depth across cells while preserving relative expression differences.

The scale factor of 10,000 (counts per ten thousand, CP10K) is conventional. It does not affect downstream results if PCA is used, because PCA is scale-invariant. It matters only when comparing raw normalized values directly.

```{r norm_rna}
# [BLOCK 3 | Step 3.1.1 | RNA LogNormalize]
sobj_filt <- NormalizeData(
  sobj_filt,
  normalization.method = "LogNormalize",
  scale.factor         = 10000
)
cat("RNA normalized. Layer 'data' now populated.\n")
cat("Layers present:", paste(Layers(sobj_filt[["RNA"]]), collapse = ", "), "\n")

# Verify: log-normalized values should be non-negative
data_layer <- LayerData(sobj_filt, assay = "RNA", layer = "data")
cat("Any negative values in RNA data layer:", any(data_layer < 0), "\n")
cat("Max value in RNA data layer          :",
    round(max(data_layer@x), 2), "\n")
```

### 3.1.2 \| ADT normalization: the wrong margin first

ADT normalization uses CLR (Centered Log-Ratio). The `margin` argument determines the direction of normalization: - `margin = 1`: normalizes each protein across all cells (makes row means \~0) - `margin = 2`: normalizes each cell across all proteins (makes column means \~0)

For ADT, `margin = 2` is correct. It removes cell-to-cell technical variation in total antibody capture efficiency, analogous to library size normalization in RNA. `margin = 1` is wrong for ADT, but it runs without any error or warning.

```{r norm_adt_wrong}
# [BLOCK 3 | Step 3.1.2 | ADT CLR with margin=1 -- WRONG, no error produced]
DefaultAssay(sobj_filt) <- "ADT"

sobj_filt <- NormalizeData(
  sobj_filt,
  normalization.method = "CLR",
  margin               = 1  # WRONG for ADT
)

# Diagnostic: with margin=1, row means (per protein) will be ~0
adt_m1 <- LayerData(sobj_filt, layer = "data")
cat("Row means after margin=1 (per protein) -- all ~0, signature of wrong margin:\n")
print(round(rowMeans(adt_m1), 4))
cat("\nColumn means after margin=1 (per cell) -- these are NOT ~0:\n")
print(round(colMeans(adt_m1)[1:8], 4))

DefaultAssay(sobj_filt) <- "RNA"
```

The row means are all near zero. That is the fingerprint of `margin=1`: each protein was independently centered across cells, removing biological variation. The data layer is now useless for any downstream protein analysis. We will see the consequence in Block 4 when CD4 gating fails.

```{r norm_adt_correct}
# [BLOCK 3 | Step 3.1.2 | ADT CLR with margin=2 -- correct]
DefaultAssay(sobj_filt) <- "ADT"

sobj_filt <- NormalizeData(
  sobj_filt,
  normalization.method = "CLR",
  margin               = 2  # CORRECT: normalizes each cell across proteins
)

adt_m2 <- LayerData(sobj_filt, layer = "data")
cat("Column means after margin=2 (per cell) -- now ~0, correct:\n")
print(round(colMeans(adt_m2)[1:8], 4))
cat("\nRow means (per protein) -- NOT ~0, biological variation preserved:\n")
print(round(rowMeans(adt_m2), 4))

DefaultAssay(sobj_filt) <- "RNA"
cat("\nADT normalized correctly.\n")
```

> **Survival moment:** `margin=1` runs silently. It is not detectable from the pipeline output unless you check the row/column means of the normalized ADT layer. In published analyses, this error has been found post-publication. Always verify: if row means are all \~0, re-run with `margin=2`.

## Step 3.2 \| Feature selection, scaling, and PCA

### 3.2.1 \| Highly variable genes

```{r hvgs}
# [BLOCK 3 | Step 3.2.1 | Highly variable gene selection]
sobj_filt <- FindVariableFeatures(
  sobj_filt,
  selection.method = "vst",
  nfeatures        = 2000
)

top10 <- head(VariableFeatures(sobj_filt), 10)
cat("Top 10 most variable genes:\n")
print(top10)
cat("\nTotal variable features selected:", length(VariableFeatures(sobj_filt)), "\n")

# Challenge: what fraction of all genes are selected as variable?
all_genes <- nrow(sobj_filt[["RNA"]])
n_var     <- length(VariableFeatures(sobj_filt))
cat("Variable fraction:", round(n_var / all_genes * 100, 1), "% of all genes\n")
```

### 3.2.2 \| Scaling and PCA

```{r scale_pca}
# [BLOCK 3 | Step 3.2.2 | Scale and run PCA]
# vars.to.regress removes the linear effect of percent.mt from each gene
# before scaling. This reduces the influence of cell stress/quality on
# the principal components.
sobj_filt <- ScaleData(
  sobj_filt,
  vars.to.regress = "percent.mt",
  verbose         = FALSE
)

sobj_filt <- RunPCA(sobj_filt, npcs = 30, verbose = FALSE)

# Inspect the top loadings of PC1 and PC2
cat("Top genes loading on PC1:\n")
print(head(sobj_filt@reductions$pca@feature.loadings[
  order(abs(sobj_filt@reductions$pca@feature.loadings[,1]),
        decreasing = TRUE), 1], 10))
```

```{r elbow, fig.height=4}
# [BLOCK 3 | Step 3.2.2 | Elbow plot to select number of PCs]
# The elbow is where adding more PCs explains little additional variance.
# Use this to set dims in FindNeighbors and RunUMAP.
ElbowPlot(sobj_filt, ndims = 30) +
  geom_vline(xintercept = 20, linetype = "dashed", color = "#991b1b") +
  annotate("text", x = 21, y = 2.5, label = "~20 PCs", hjust = 0,
           color = "#991b1b") +
  ggtitle("Elbow Plot: variance explained per PC")
```

## Step 3.3 \| Clustering and UMAP

### 3.3.1 \| Inconsistent dims between RunPCA and FindNeighbors

```{r neighbors_wrong, error=TRUE}
# [BLOCK 3 | Step 3.3.1 | FindNeighbors with dims > npcs -- ERROR]
# npcs=30 was used in RunPCA. What happens if we request 50 dims?
sobj_filt <- FindNeighbors(sobj_filt, dims = 1:50, verbose = FALSE)
```

The error occurs because the PCA object contains only 30 components. `dims = 1:50` requests components that do not exist. This is common when adjusting `npcs` in RunPCA without updating downstream calls.

```{r neighbors_correct}
# [BLOCK 3 | Step 3.3.1 | Correct: dims consistent with RunPCA]
sobj_filt <- FindNeighbors(sobj_filt, dims = 1:20, verbose = FALSE)
```

### 3.3.2 \| Resolution sensitivity

The `resolution` parameter in `FindClusters` is one of the most consequential but arbitrary choices in the workflow. Higher resolution = more, smaller clusters. There is no single correct value. The choice should be validated against known biology (do canonical marker genes match cluster assignments?).

```{r clustering_resolution, fig.width=12, fig.height=4}
# [BLOCK 3 | Step 3.3.2 | Resolution sensitivity]
# How much does the clustering change between resolution 0.2 and 1.2?

res_values <- c(0.2, 0.5, 1.2)
n_clusters <- sapply(res_values, function(r) {
  tmp <- FindClusters(sobj_filt, resolution = r, verbose = FALSE)
  length(unique(tmp$seurat_clusters))
})

cat("Resolution vs number of clusters:\n")
for (i in seq_along(res_values)) {
  cat("  resolution =", res_values[i], "->", n_clusters[i], "clusters\n")
}
cat("\nUsing resolution = 0.5 for the rest of the analysis.\n")
```

```{r final_clustering}
# [BLOCK 3 | Step 3.3.2 | Final clustering at resolution 0.5]
sobj_filt <- FindClusters(sobj_filt, resolution = 0.5, verbose = FALSE)
cat("Clusters found:", length(unique(sobj_filt$seurat_clusters)), "\n")
print(table(sobj_filt$seurat_clusters))
```

### 3.3.3 \| UMAP and condition overlay

```{r umap, fig.width=12, fig.height=5}
# [BLOCK 3 | Step 3.3.3 | UMAP with cluster and donor visualization]
# Three questions to answer immediately after generating the UMAP:
# 1. Do cells cluster by biology or by donor? (batch effect check)
# 2. Do canonical marker genes map to expected clusters?
# 3. Does condition (Healthy vs COVID-19) show any spatial structure?
set.seed(42)
sobj_filt <- RunUMAP(sobj_filt, dims = 1:20, verbose = FALSE)

p1 <- DimPlot(sobj_filt, group.by = "seurat_clusters",
              label = TRUE, label.size = 4) +
  NoLegend() + ggtitle("Clusters (resolution = 0.5)")

p2 <- DimPlot(sobj_filt, group.by = "orig.ident") +
  ggtitle("By donor: batch effect?") +
  theme(legend.text = element_text(size = 7))

p1 | p2
```

```{r umap_condition, fig.width=12, fig.height=5}
# [BLOCK 3 | Step 3.3.3 | Condition and severity on UMAP]
p_cond <- DimPlot(sobj_filt, group.by = "condition",
                  cols = c("Healthy" = "#0d7377", "COVID19" = "#991b1b"),
                  pt.size = 0.5) +
  ggtitle("Healthy vs COVID-19")

p_sev <- DimPlot(sobj_filt, group.by = "severity",
                 pt.size = 0.5) +
  ggtitle("COVID-19 severity") +
  theme(legend.text = element_text(size = 8))

p_cond | p_sev
```

```{r markers_rna, fig.height=8}
# [BLOCK 3 | Step 3.3.3 | Canonical PBMC markers on UMAP]
# Match marker expression to cluster locations before formal annotation
FeaturePlot(sobj_filt,
            features   = c("CD3E", "CD19", "CD14", "NCAM1", "IL7R", "FCGR3A"),
            ncol       = 3,
            min.cutoff = "q05")
```

```{r dot_plot, fig.height=5}
# [BLOCK 3 | Step 3.3.3 | Dot plot: expression level and frequency by cluster]
DotPlot(sobj_filt,
        features = c("CD3E", "CD4", "CD8A", "CD19", "MS4A1",
                     "CD14", "FCGR3A", "NCAM1", "PPBP"),
        group.by = "seurat_clusters") +
  RotatedAxis() +
  ggtitle("Marker expression by cluster")
```

```{r prop_condition, fig.height=4}
# [BLOCK 3 | Step 3.3.3 | Cluster proportion by condition]
# This is a preliminary result. It must be validated after annotation.
props <- sobj_filt@meta.data %>%
  group_by(condition, seurat_clusters) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(condition) %>%
  mutate(prop = n / sum(n))

ggplot(props, aes(x = condition, y = prop, fill = seurat_clusters)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_y_continuous(labels = percent_format()) +
  labs(title    = "Cluster proportions: Healthy vs COVID-19",
       subtitle = "Check per-donor before interpreting any group difference",
       x = "Condition", y = "Proportion", fill = "Cluster") +
  theme_classic(base_size = 12)
```

> **Survival moment:** A cluster that is visually larger in COVID-19 than Healthy might be driven entirely by one donor with unusual composition. Never report a condition effect from aggregated data without first checking `DimPlot(group.by = "orig.ident")` and confirming the pattern is consistent across all donors within each group.

```{r checkpoint}
# [BLOCK 3 | Checkpoint save]
saveRDS(sobj_filt, "outputs/sobj_preprocessed.rds")
cat("Checkpoint saved: outputs/sobj_preprocessed.rds\n")
cat("To reload: sobj_filt <- readRDS('outputs/sobj_preprocessed.rds')\n")
```

------------------------------------------------------------------------

# BREAK: 15 minutes \| 1:55 to 2:10

Look at the UMAP colored by condition. Does any cluster appear exclusively or predominantly in COVID-19 donors? Look at the DotPlot: which clusters could be monocytes based on CD14 and FCGR3A expression? After the break, we add the protein layer to test these interpretations.

------------------------------------------------------------------------

------------------------------------------------------------------------

# ══════════════════════════════════════════

# BLOCK 4 \| CITE-seq Integrative Analysis

# ══════════════════════════════════════════

*2:10 to 3:10 \| 60 min*

------------------------------------------------------------------------

## Step 4.1 \| Cell type annotation with SingleR

SingleR assigns a cell type label by correlating each cell's normalized expression profile against a reference dataset of known cell types. It uses Spearman correlation across the top marker genes for each reference label, then selects the label with the highest correlation.

The main limitation: any cell type absent from the reference will be assigned to the most similar reference type. Low-confidence assignments are flagged by a pruning step (`pruned.labels`). Cells with NA in pruned labels had too low a score difference to be reliably assigned.

```{r singler_run}
# [BLOCK 4 | Step 4.1 | Run SingleR against Human Primary Cell Atlas]
cat("Loading reference (Human Primary Cell Atlas)...\n")
ref     <- celldex::HumanPrimaryCellAtlasData()
rna_mat <- LayerData(sobj_filt, assay = "RNA", layer = "data")

cat("Running SingleR...\n")
singler_res <- SingleR(
  test    = rna_mat,
  ref     = ref,
  labels  = ref$label.main,
  BPPARAM = BiocParallel::SerialParam()
)

cat("\nCell type distribution:\n")
print(sort(table(singler_res$labels), decreasing = TRUE))

cat("\nLow-confidence cells (NA in pruned labels):",
    sum(is.na(singler_res$pruned.labels)), "\n")
```

### 4.1.1 \| Plotting before adding labels to the object

```{r singler_plot_early, error=TRUE}
# [BLOCK 4 | Step 4.1.1 | Plotting singler_label before adding it -- ERROR]
# singler_res is a separate object. Labels must be transferred to sobj_filt first.
DimPlot(sobj_filt, group.by = "singler_label")
```

`singler_label` does not exist in `sobj_filt@meta.data` yet. `singler_res` is a `DataFrame` (Bioconductor class). The labels must be explicitly added to the Seurat metadata.

```{r singler_add_labels}
# [BLOCK 4 | Step 4.1.1 | Add SingleR labels to Seurat object]
sobj_filt$singler_label  <- singler_res$labels
sobj_filt$singler_pruned <- singler_res$pruned.labels

# Verify
cat("singler_label in metadata:", "singler_label" %in% colnames(sobj_filt@meta.data), "\n")
cat("Distribution:\n")
print(sort(table(sobj_filt$singler_label), decreasing = TRUE))
```

```{r singler_umap, fig.width=12, fig.height=5}
# [BLOCK 4 | Step 4.1.1 | Annotated UMAP + condition overlay]
p_ann <- DimPlot(sobj_filt, group.by = "singler_label",
                 label = TRUE, label.size = 3, repel = TRUE) +
  ggtitle("SingleR annotation") + NoLegend()

p_cond <- DimPlot(sobj_filt, group.by = "condition",
                  cols = c("Healthy" = "#0d7377", "COVID19" = "#991b1b"),
                  pt.size = 0.4) +
  ggtitle("Condition")

p_ann | p_cond
```

```{r singler_qc_scores, fig.height=5}
# [BLOCK 4 | Step 4.1.1 | Score heatmap: annotation confidence]
# Cells with similar scores across multiple types are ambiguous.
# Wide spread = high confidence. Narrow spread = low confidence.
plotScoreHeatmap(singler_res,
                 main = "SingleR annotation scores (wider spread = more confident)")
```

## Step 4.2 \| RNA vs protein: the CITE-seq contrast

### 4.2.1 \| The DefaultAssay problem

```{r adt_wrong_assay, fig.width=12, fig.height=5}
# [BLOCK 4 | Step 4.2.1 | Plotting ADT without switching DefaultAssay]
# What happens when we forget to set DefaultAssay("ADT") before FeaturePlot?
cat("Current DefaultAssay:", DefaultAssay(sobj_filt), "\n")

# This will plot RNA, not ADT -- no error, just the wrong modality
p_forgotten <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("CD4: DefaultAssay = RNA (forgot to switch)") +
  scale_color_gradient(low = "lightgrey", high = "#374151")

print(p_forgotten)
```

This plot shows CD4 RNA. It looks dim and patchy because of dropout. The problem is entirely silent: no error, no warning, just a biologically misleading visualization. If you labeled this as "CD4 protein expression" in a figure, it would be factually incorrect.

```{r rna_vs_adt_contrast, fig.width=12, fig.height=5}
# [BLOCK 4 | Step 4.2.1 | Correct comparison: RNA vs ADT for CD4]
DefaultAssay(sobj_filt) <- "RNA"
p_rna <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("CD4 via RNA\n(dropout: many CD4+ cells read as zero)") +
  scale_color_gradient(low = "lightgrey", high = "#991b1b")

DefaultAssay(sobj_filt) <- "ADT"
p_adt <- FeaturePlot(sobj_filt, "CD4", min.cutoff = "q05") +
  ggtitle("CD4 via ADT protein\n(bimodal, reliable)") +
  scale_color_gradient(low = "lightgrey", high = "#0d7377")

p_rna | p_adt
DefaultAssay(sobj_filt) <- "RNA"
```

> **This is the core argument for CITE-seq.** CD4 RNA shows extensive dropout in true CD4+ T cells. CD4 protein from ADT is clean and bimodal. If you annotate T cell subsets using RNA alone, you will misclassify a fraction of CD4+ cells because their CD4 RNA reads as zero.

```{r rna_adt_correlation}
# [BLOCK 4 | Step 4.2.2 | Quantify RNA-protein correlation per marker]
# For each ADT protein with an RNA counterpart, compute the Spearman
# correlation across all cells. Low correlation = high dropout in RNA.

adt_rna_pairs <- list(
  CD3  = "CD3E",
  CD4  = "CD4",
  CD8a = "CD8A",
  CD14 = "CD14",
  CD19 = "CD19",
  CD56 = "NCAM1"
)

DefaultAssay(sobj_filt) <- "ADT"
adt_vals <- t(LayerData(sobj_filt, layer = "data"))
DefaultAssay(sobj_filt) <- "RNA"
rna_vals <- t(LayerData(sobj_filt, layer = "data"))

cat("Spearman correlation (RNA vs ADT) per marker:\n")
cat(sprintf("  %-8s %s\n", "Marker", "Spearman rho"))
cat(sprintf("  %-8s %s\n", "------", "------------"))

for (adt_name in names(adt_rna_pairs)) {
  rna_gene <- adt_rna_pairs[[adt_name]]
  if (rna_gene %in% colnames(rna_vals) &&
      adt_name %in% colnames(adt_vals)) {
    rho <- cor(rna_vals[, rna_gene],
               adt_vals[, adt_name],
               method = "spearman")
    cat(sprintf("  %-8s %.3f\n", adt_name, rho))
  }
}
DefaultAssay(sobj_filt) <- "RNA"
```

> **Challenge 7:** Which markers show the lowest RNA-protein correlation? Are these the markers where dropout would most severely affect annotation? What biological factors (expression level, transcript stability) might explain why some markers have higher correlation than others?

```{r dropout_quantification}
# [BLOCK 4 | Step 4.2.2 | Quantify dropout: cells positive by ADT but zero by RNA]
# For each marker, find cells that are clearly protein-positive (ADT > threshold)
# but have zero RNA counts. These are the cells that would be misannotated
# by RNA-only analysis.

DefaultAssay(sobj_filt) <- "ADT"
adt_data <- LayerData(sobj_filt, layer = "data")
DefaultAssay(sobj_filt) <- "RNA"
rna_counts <- LayerData(sobj_filt, layer = "counts")

cat("Dropout analysis: cells positive by ADT but zero by RNA\n")
cat(sprintf("  %-8s %-12s %-12s %s\n",
            "Marker", "ADT+ cells", "RNA==0 among ADT+", "Dropout rate"))
cat(sprintf("  %-8s %-12s %-12s %s\n",
            "------", "----------", "-----------------", "------------"))

marker_pairs <- list(CD4 = "CD4", CD14 = "CD14", CD19 = "CD19")
for (adt_name in names(marker_pairs)) {
  rna_gene <- marker_pairs[[adt_name]]
  if (rna_gene %in% rownames(rna_counts) && adt_name %in% rownames(adt_data)) {
    adt_pos  <- which(adt_data[adt_name, ] > 1.0)
    rna_zero <- sum(rna_counts[rna_gene, adt_pos] == 0)
    pct      <- round(rna_zero / length(adt_pos) * 100, 1)
    cat(sprintf("  %-8s %-12d %-12d %s%%\n",
                adt_name, length(adt_pos), rna_zero, pct))
  }
}
```

## Step 4.3 \| Digital gating

Digital gating replicates the biaxial scatter plots from flow cytometry using ADT data. It allows computational cell classification with the same logic immunologists use at the bench.

```{r digital_gating, fig.height=5}
# [BLOCK 4 | Step 4.3 | CD4 vs CD8a digital gating]
DefaultAssay(sobj_filt) <- "ADT"

color_by <- if ("singler_label" %in% colnames(sobj_filt@meta.data)) {
  "singler_label"
} else {
  "cell_type"
}

gate_data <- FetchData(sobj_filt, vars = c("CD4", "CD8a", color_by))
names(gate_data)[3] <- "label"

ggplot(gate_data, aes(CD4, CD8a, color = label)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_density_2d(color = "grey50", linewidth = 0.3) +
  geom_vline(xintercept = 1.5, linetype = "dashed", color = "#991b1b") +
  geom_hline(yintercept = 1.5, linetype = "dashed", color = "#991b1b") +
  annotate("text", x = 3.5, y = 0.4, label = "CD4+ T cells", size = 3.5) +
  annotate("text", x = 0.4, y = 3.5, label = "CD8+ T cells", size = 3.5) +
  labs(title    = "Digital gating: CD4 vs CD8a (ADT protein)",
       subtitle = "Only possible with ADT; RNA dropout makes this scatter uninformative",
       x = "CD4 (CLR normalized)", y = "CD8a (CLR normalized)",
       color = NULL) +
  theme_classic(base_size = 13) +
  theme(legend.text = element_text(size = 7))

DefaultAssay(sobj_filt) <- "RNA"
```

## Step 4.4 \| WNN: combining RNA and protein

Weighted Nearest Neighbor integration learns the relative contribution of each modality per cell. A cell where RNA is informative (high complexity, many genes detected) receives more RNA weight. A cell where protein is more discriminating (e.g., ambiguous T cell subtype) receives more ADT weight.

WNN requires a PCA for each modality separately before integration.

### 4.4.1 \| Running WNN without the ADT PCA first

```{r wnn_missing_reduction, error=TRUE}
# [BLOCK 4 | Step 4.4.1 | WNN without ADT PCA -- ERROR]
# RunPCA was only called on RNA in Block 3. 'pca.adt' does not exist yet.
sobj_filt <- FindMultiModalNeighbors(
  sobj_filt,
  reduction.list       = list("pca", "pca.adt"),
  dims.list            = list(1:20, 1:15),
  modality.weight.name = "RNA.weight",
  verbose              = FALSE
)
```

The reduction `pca.adt` was never computed. WNN requires a separate PCA for each modality. We must scale the ADT data and run PCA on it first.

```{r wnn_adt_pca}
# [BLOCK 4 | Step 4.4.1 | Compute ADT PCA first, then run WNN]
# ADT has only 24 proteins. FindVariableFeatures is not needed.
# Both ScaleData and RunPCA require an explicit features list when
# no VariableFeatures exist for the assay.
DefaultAssay(sobj_filt) <- "ADT"

adt_features <- rownames(sobj_filt[["ADT"]])  # all 24 proteins

sobj_filt <- ScaleData(
  sobj_filt,
  features = adt_features,
  verbose  = FALSE
)

sobj_filt <- RunPCA(
  sobj_filt,
  features       = adt_features,
  npcs           = 20,
  reduction.name = "pca.adt",
  reduction.key  = "pcaADT_",
  verbose        = FALSE
)

cat("Reductions now available:", paste(Reductions(sobj_filt), collapse = ", "), "\n")
```

```{r wnn_compute}
# [BLOCK 4 | Step 4.4.1 | WNN integration]
sobj_filt <- FindMultiModalNeighbors(
  sobj_filt,
  reduction.list       = list("pca", "pca.adt"),
  dims.list            = list(1:20, 1:15),
  modality.weight.name = "RNA.weight",
  verbose              = FALSE
)

sobj_filt <- RunUMAP(
  sobj_filt,
  nn.name        = "weighted.nn",
  reduction.name = "wnn.umap",
  reduction.key  = "wnnUMAP_",
  verbose        = FALSE
)

DefaultAssay(sobj_filt) <- "RNA"
cat("WNN complete. New reduction: wnn.umap\n")
```

```{r wnn_compare, fig.width=12, fig.height=5}
# [BLOCK 4 | Step 4.4.1 | RNA-only UMAP vs WNN UMAP]
p_rna <- DimPlot(sobj_filt, reduction = "umap",
                 group.by = "singler_label",
                 label = TRUE, label.size = 2.5, repel = TRUE) +
  ggtitle("RNA-only UMAP") + NoLegend()

p_wnn <- DimPlot(sobj_filt, reduction = "wnn.umap",
                 group.by = "singler_label",
                 label = TRUE, label.size = 2.5, repel = TRUE) +
  ggtitle("WNN UMAP (RNA + ADT)") + NoLegend()

p_rna | p_wnn
```

```{r modality_weights, fig.height=4}
# [BLOCK 4 | Step 4.4.1 | Modality weights: which cells rely more on ADT?]
# RNA.weight close to 1 = cell is better characterized by RNA
# RNA.weight close to 0 = cell is better characterized by ADT

ggplot(sobj_filt@meta.data, aes(x = RNA.weight, fill = singler_label)) +
  geom_histogram(bins = 40, alpha = 0.8) +
  facet_wrap(~singler_label, scales = "free_y") +
  labs(title    = "Per-cell RNA modality weight from WNN",
       subtitle = "Low weight = ADT more informative for that cell",
       x = "RNA weight (1 = fully RNA, 0 = fully ADT)") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none",
        strip.text      = element_text(size = 7))
```

## Step 4.5 \| Condition comparison: Healthy vs COVID-19

### 4.5.1 \| Group-level plot without per-donor validation

```{r covid_group_plot, fig.height=4}
# [BLOCK 4 | Step 4.5.1 | Condition comparison -- group level only]
# This plot looks informative. But with 3 Healthy and 4 COVID-19 donors,
# it can be completely driven by one outlier donor.
group_props <- sobj_filt@meta.data %>%
  filter(!is.na(singler_label), !is.na(condition)) %>%
  group_by(condition, singler_label) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(condition) %>%
  mutate(prop = n / sum(n))

ggplot(group_props, aes(x = condition, y = prop, fill = singler_label)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_y_continuous(labels = percent_format()) +
  labs(title    = "Cell type proportions by condition",
       subtitle = "Looks clear -- but is it driven by one donor?",
       x = "Condition", y = "Proportion", fill = "Cell type") +
  theme_classic(base_size = 12) +
  theme(legend.text = element_text(size = 8))
```

```{r covid_per_donor, fig.height=5}
# [BLOCK 4 | Step 4.5.1 | Per-donor validation of group difference]
donor_props <- sobj_filt@meta.data %>%
  filter(!is.na(singler_label)) %>%
  group_by(donor_id, condition, singler_label) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(donor_id) %>%
  mutate(prop = n / sum(n))

ggplot(donor_props, aes(x = donor_id, y = prop, fill = singler_label)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_y_continuous(labels = percent_format()) +
  facet_wrap(~condition, scales = "free_x") +
  labs(title    = "Cell type proportions per donor",
       subtitle = "Consistent pattern across donors within each group = robust finding",
       x = NULL, y = "Proportion (%)", fill = "Cell type") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.text = element_text(size = 7),
        strip.text  = element_text(face = "bold"))
```

> **Survival moment:** If the group-level difference is driven by a single donor, the finding is not generalizable. With n=3 per group, you cannot perform a standard statistical test. The minimum defensible design for condition-level claims is 4-5 donors per group with at least 1,000 cells per donor. Always report n per group in any figure legend.

```{r save_annotated}
saveRDS(sobj_filt, "outputs/sobj_annotated.rds")
cat("Annotated object saved: outputs/sobj_annotated.rds\n")
```

------------------------------------------------------------------------

------------------------------------------------------------------------

> **Discussion pause**
>
> - The WNN modality weights show that some cell types rely more on ADT than RNA. Which cell types had the lowest RNA weight? Why does that make sense?
> - The dropout rate for CD4 RNA was calculated above. What would a naive RNA-only analysis conclude about those cells?
> - You see monocytes are expanded in COVID-19. How would you test whether this is a statistically significant finding given n=3 Healthy donors?

------------------------------------------------------------------------

------------------------------------------------------------------------

# ══════════════════════════════════════════

# BLOCK 5 \| Troubleshooting Scenarios

# ══════════════════════════════════════════

*3:10 to 4:10 \| 60 min \| groups of 3 to 4*

------------------------------------------------------------------------

> **Instructions** The instructor has posted `sobj_broken.rds` in Google Classroom right now. Download it and save to `data/sobj_broken.rds`. For each scenario: run the diagnostic chunks, write your group hypothesis as a comment in the code, then expand the solution. The hypothesis step is mandatory.

```{r load_broken}
# [BLOCK 5 | Load broken object]
sobj_broken <- readRDS("data/sobj_broken.rds")
print(sobj_broken)
head(sobj_broken@meta.data, 3)
```

------------------------------------------------------------------------

## Scenario 1 \| Wrong default assay (15 min)

**Context:** You received this object from a collaborator. You run the standard RNA preprocessing workflow. Everything executes without errors but the results are completely wrong: FindVariableFeatures returns fewer features than expected, PCA explains almost all variance in PC1, and the UMAP is a single blob.

```{r s1_diag1}
# [BLOCK 5 | Scenario 1 | Diagnostic]
cat("Default assay        :", DefaultAssay(sobj_broken), "\n")
cat("Features active assay:", nrow(sobj_broken), "\n")
cat("All assays           :", paste(Assays(sobj_broken), collapse = ", "), "\n")
```

```{r s1_diag2}
# What does FindVariableFeatures return on this object?
test_hvg <- FindVariableFeatures(sobj_broken, nfeatures = 2000, verbose = FALSE)
cat("Variable features found:", length(VariableFeatures(test_hvg)), "\n")
cat("Feature names:\n")
print(head(VariableFeatures(test_hvg), 15))
```

```{r s1_diag3}
# Compare RNA and ADT feature counts
cat("Features in RNA assay:", nrow(sobj_broken[["RNA"]]), "\n")
cat("Features in ADT assay:", nrow(sobj_broken[["ADT"]]), "\n")
```

> 📝 **Write your group hypothesis before expanding the solution.** What is the DefaultAssay? Why does FindVariableFeatures return so few features? What would happen if you ran RunPCA and RunUMAP on this object without fixing it? How would you detect this problem in a published Seurat object you downloaded?

```{r s1_fix}
# [BLOCK 5 | Scenario 1 | Solution]
cat("Before fix:", DefaultAssay(sobj_broken), "\n")

# The DefaultAssay is set to "ADT" (24 proteins).
# All RNA functions are running silently on 24 protein features instead
# of 500 RNA genes. FindVariableFeatures returns 24 features because that
# is all that exists in the active assay.

DefaultAssay(sobj_broken) <- "RNA"
cat("After fix :", DefaultAssay(sobj_broken), "\n")
cat("Features  :", nrow(sobj_broken), "\n")

test_fixed <- FindVariableFeatures(sobj_broken, nfeatures = 2000, verbose = FALSE)
cat("Variable features now:", length(VariableFeatures(test_fixed)), "\n")
```

**Lessons learned:**

1.  `DefaultAssay()` is the first line to run on any received object.
2.  This error is completely silent. No warning. No error. Just wrong results.
3.  `FindVariableFeatures`, `ScaleData`, `RunPCA`, `FindMarkers` all operate on the active assay. A PCA run on 24 ADT features is not a transcriptomic PCA.
4.  Always check `DefaultAssay()` after any assay switch to confirm it reset.

------------------------------------------------------------------------

## Scenario 2 \| Metadata inconsistencies (20 min)

**Context:** This object was assembled from samples processed at multiple sites. Grouping by `condition` and `donor_id` produces unexpected results. Some samples are missing from plots and certain donors appear duplicated.

```{r s2_diag1}
# [BLOCK 5 | Scenario 2 | Diagnostic: metadata audit]
cat("Metadata columns:\n")
print(colnames(sobj_broken@meta.data))

cat("\nNAs per column:\n")
print(colSums(is.na(sobj_broken@meta.data)))
```

```{r s2_diag2}
# Inspect the problematic columns
cat("Unique donor_id values:\n")
print(sort(unique(sobj_broken$donor_id)))

cat("\nCondition table (including NAs):\n")
print(table(sobj_broken$condition, useNA = "always"))

cat("\nCell type NAs:", sum(is.na(sobj_broken$cell_type)), "\n")
```

```{r s2_diag3}
# What does a DimPlot by condition look like?
DimPlot(sobj_broken, group.by = "condition") +
  ggtitle("Condition plot: how many groups appear?")
```

> 📝 **Write your group hypothesis before expanding the solution.** How many distinct problems did you find? What would the biological consequence be if each went unfixed? For example: if condition labels are inconsistent, what happens to a Healthy vs COVID-19 comparison?

<details>

<summary>Reveal solution</summary>

```{r s2_fix}
# [BLOCK 5 | Scenario 2 | Solution]
sobj_meta <- sobj_broken

# PROBLEM 1: donor_id has three different formats for the same donors
# e.g. "Donor01", "donor1", "DONOR01" all mean the same donor
cat("=== FIX 1: donor_id ===\n")
sobj_meta$donor_id <- toupper(gsub("[^A-Z0-9a-z]", "", sobj_meta$donor_id))
print(table(sobj_meta$donor_id))

# PROBLEM 2: condition has 4 variants of 2 values
# "COVID19", "covid19", "Covid-19", "HC", NA
cat("\n=== FIX 2: condition ===\n")
sobj_meta$condition_clean <- dplyr::case_when(
  grepl("covid", sobj_meta$condition, ignore.case = TRUE) ~ "COVID-19",
  grepl("healthy|HC|control", sobj_meta$condition,
        ignore.case = TRUE)                               ~ "Healthy",
  TRUE                                                    ~ NA_character_
)
print(table(sobj_meta$condition_clean, useNA = "always"))

# PROBLEM 3: 200 cells have NA in cell_type
cat("\n=== FIX 3: cell_type NAs ===\n")
cat("NAs before:", sum(is.na(sobj_meta$cell_type)), "\n")
sobj_meta$cell_type[is.na(sobj_meta$cell_type)] <- "Unknown"
cat("NAs after :", sum(is.na(sobj_meta$cell_type)), "\n")
```

**Lessons learned:**

1.  `table(col, useNA = "always")` and `str(meta.data)` before every analysis.
2.  Multi-site data almost always has formatting inconsistencies.
3.  NA in condition: that cell is excluded from any Healthy vs COVID-19 comparison. With 30 NAs, you are silently dropping 1% of cells from all group comparisons.
4.  Case-sensitivity matters: `"COVID19" == "covid19"` is `FALSE` in R.

</details>

------------------------------------------------------------------------

## Scenario 3 \| Normalization failure and batch effect (25 min)

**Context:** The UMAP shows clear separation by donor rather than by cell type. Additionally, the ADT data appears distorted. Identify both problems and fix them.

```{r s3_diag1, fig.height=4}
# [BLOCK 5 | Scenario 3 | Diagnostic: batch effect in UMAP]
DimPlot(sobj_broken, group.by = "orig.ident") +
  ggtitle("Donors separate in UMAP: batch effect or biology?")
```

```{r s3_diag2}
# [BLOCK 5 | Scenario 3 | Diagnostic: ADT normalization direction]
DefaultAssay(sobj_broken) <- "ADT"
adt_data <- LayerData(sobj_broken, layer = "data")

cat("Row means (per protein) -- near 0 means margin=1 was used (wrong):\n")
print(round(rowMeans(adt_data), 4))

cat("\nColumn means (per cell, first 10) -- near 0 means margin=2 (correct):\n")
print(round(colMeans(adt_data)[1:10], 4))

DefaultAssay(sobj_broken) <- "RNA"
```

```{r s3_diag3}
# [BLOCK 5 | Scenario 3 | Diagnostic: RNA data layer integrity]
rna_data <- LayerData(sobj_broken, assay = "RNA", layer = "data")
cat("Any negative values in RNA data layer?",
    any(rna_data < 0), "\n")
cat("(TRUE would indicate SCTransform residuals stored as 'data')\n")
cat("\nRange of non-zero values in RNA data:\n")
print(summary(rna_data@x))
```

> 📝 **Write your group hypothesis before expanding the solution.** How did you determine which margin was used? What is the consequence of the batch effect for any condition-level comparison? If you published the original UMAP as a figure, what scientific claim would be invalidated?

<details>

<summary>Reveal solution</summary>

```{r s3_fix}
# [BLOCK 5 | Scenario 3 | Solution: re-normalize and apply Harmony]
sobj_renorm <- sobj_broken

# STEP 1: Re-normalize RNA from counts (counts layer is untouched)
DefaultAssay(sobj_renorm) <- "RNA"
sobj_renorm <- NormalizeData(sobj_renorm,
  normalization.method = "LogNormalize",
  scale.factor         = 10000,
  verbose              = FALSE)
cat("RNA re-normalized from counts.\n")

# STEP 2: Re-normalize ADT with correct margin
DefaultAssay(sobj_renorm) <- "ADT"
sobj_renorm <- NormalizeData(sobj_renorm,
  normalization.method = "CLR",
  margin               = 2,
  verbose              = FALSE)
cat("ADT re-normalized (CLR, margin=2).\n")

# STEP 3: Rerun preprocessing
DefaultAssay(sobj_renorm) <- "RNA"
sobj_renorm <- FindVariableFeatures(sobj_renorm, verbose = FALSE)
sobj_renorm <- ScaleData(sobj_renorm, verbose = FALSE)
sobj_renorm <- RunPCA(sobj_renorm, npcs = 30, verbose = FALSE)

# STEP 4: Harmony batch correction
sobj_renorm <- RunHarmony(
  sobj_renorm,
  group.by.vars  = "orig.ident",
  reduction      = "pca",
  reduction.save = "harmony",
  verbose        = FALSE
)

sobj_renorm <- RunUMAP(sobj_renorm,
  reduction      = "harmony",
  dims           = 1:20,
  reduction.name = "umap.harmony",
  verbose        = FALSE)

cat("Harmony applied.\n")
```

```{r s3_compare, fig.width=12, fig.height=4}
p_before <- DimPlot(sobj_renorm, reduction = "umap",
                    group.by = "orig.ident") +
  ggtitle("Before Harmony") + theme(legend.position = "none")

p_after  <- DimPlot(sobj_renorm, reduction = "umap.harmony",
                    group.by = "orig.ident") +
  ggtitle("After Harmony") + theme(legend.position = "none")

p_before | p_after
```

**Lessons learned:**

1.  Row means \~0 in ADT data is the diagnostic signature of `margin=1`. Wrong.
2.  Column means \~0 in ADT data is the signature of `margin=2`. Correct.
3.  The counts layer is the ground truth. Re-normalize from it; never overwrite it.
4.  Donor separation in UMAP = batch effect. Run `DimPlot(group.by="orig.ident")` before interpreting any UMAP biologically.
5.  Harmony corrects for donor-level batch effects while preserving biological variation. It requires at least 3 donors per group to be reliable.

</details>

------------------------------------------------------------------------

------------------------------------------------------------------------

# ══════════════════════════════════════════

# BLOCK 6 \| Wrap-up

# ══════════════════════════════════════════

*4:10 to 4:50 \| 40 min*

------------------------------------------------------------------------

## Step 6.1 \| Biological summary: Healthy vs COVID-19

```{r final_proportions, fig.height=5}
# [BLOCK 6 | Step 6.1 | Final annotated proportions]
if ("singler_label" %in% colnames(sobj_filt@meta.data)) {
  final_props <- sobj_filt@meta.data %>%
    filter(!is.na(condition), !is.na(singler_label)) %>%
    group_by(condition, singler_label) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(condition) %>%
    mutate(prop = n / sum(n))

  ggplot(final_props, aes(condition, prop, fill = singler_label)) +
    geom_bar(stat = "identity", width = 0.65) +
    scale_y_continuous(labels = percent_format()) +
    labs(title = "Cell type proportions: final annotated result",
         x = "Condition", y = "Proportion (%)", fill = "Cell type") +
    theme_classic(base_size = 12) +
    theme(legend.text = element_text(size = 8),
          axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Step 6.2 \| Pitfalls reference table

```{r pitfalls_final, echo=FALSE}
# [BLOCK 6 | Step 6.2 | Pitfall reference]
pit <- data.frame(
  Pitfall = c(
    "UMAP interpreted as distance",
    "Per-cell differential expression",
    "Single annotation source",
    "ADT with margin=1",
    "Wrong DefaultAssay",
    "Group comparison without per-donor check",
    "Threshold copied from tutorial"
  ),
  Consequence = c(
    "Cluster proximity in UMAP has no quantitative meaning",
    "Inflated false positives by pseudoreplication",
    "Rare or novel types misclassified",
    "Protein normalization wrong; no error produced",
    "Analysis on wrong modality; no error produced",
    "Group effect driven by single donor",
    "No cells remaining after filtering"
  ),
  Prevention = c(
    "Use graph-based distance, not UMAP coordinates",
    "DESeq2 with pseudobulk aggregation per donor",
    "Validate with RNA markers + ADT + external reference",
    "Check rowMeans of ADT data layer; must NOT be ~0",
    "Check DefaultAssay() as first line on any received object",
    "Always facet or stratify by donor_id",
    "Inspect data distribution before setting any threshold"
  )
)
knitr::kable(pit, col.names = c("Pitfall", "Consequence", "Prevention"))
```

## Step 6.3 \| Session info

```{r session_info}
# [BLOCK 6 | Step 6.3 | Session info -- save with every analysis]
sessionInfo()
```

------------------------------------------------------------------------

> **You completed the Survival Course.** Every error in this notebook occurs in real published analyses, often silently. The skills practiced here: auditing objects before trusting them, verifying normalization by diagnostic checks, separating batch effects from biology, cross-validating annotation with RNA and protein, and checking per-donor consistency before any group-level claim.

*Single-Cell Survival Course \| Seurat v5 \| GSE149689 \| Wilk et al., Nature Medicine 2021*
